---
title: "Testing for normality in R"
author: "Avalon C.S. Owens, Eric R. Scott"
date: "10/26/2018"
output:
  ioslides_presentation: default
  pdf_document: default
---
## Note for next time:

- Switch `chickwts` to another more t-test appropriate dataset (maybe `sleep`)
- Curate titanic data better.  E.g. survived column should either be survival or values in it should be yes and no.
- Emphasized *why* you need to filter to split data verbally and with whiteboard, but could include this as a slide showing results of `shapiro.test()` on two separate normals with different means, and on combined data.


```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(error = TRUE)
```

## Packages for today

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(car)
skylight = read.csv("skylight.csv")
titanic = read.csv("titanic.csv")
```

## Plan for today

- Transforming your data
    + `mutate()` to add columns
    + `filter()` to select rows
- Testing your test assumptions
    + `shapiro.test()` for normality
    + `leveneTest()` for equal variance
- Normal probability plots with `ggplot2`
- Non-parametric tests (part 1 of 2)
    + `wilcox.test()`

# Transforming your data

## Transforming your data

What to do when your data don't look like you'd like...?

```{r}
skylight = read.csv("skylight.csv")
skylight$sky_type <- factor(skylight$sky_type, 
                            levels = c("skyglow",
                                      "twilight",
                                      "starlight"))
str(skylight)
```


## Transforming your data

```{r echo=FALSE}
ggplot(skylight, aes(x = wavelength, 
                     y = intensity, 
                     color = sky_type)) + 
  geom_line() + theme_dark() + 
  scale_color_manual(values=c("pink",
                              "#ccccff", 
                              "lavender"), 
                     name="night sky",
                     labels=c("skyglow", 
                              "twilight", 
                              "starlight")) +
  labs(y = expression("irradiance"~(photons~cm^{-2}~s^{-1})), x = "wavelength (nm)")
```

## Transforming your data

```{r}
ggplot(skylight, aes(x = wavelength, y = intensity, 
                     color = sky_type)) + geom_line()
```

## Transforming your data

```{r}
ggplot(skylight, aes(x = wavelength, y = log(intensity), 
                     color = sky_type)) + geom_line()
```


## Transforming your data with `mutate()`

```{r}
head(skylight)
log_skylight <- skylight %>% mutate(log_intensity = log(intensity))
```


## Transforming your data with `mutate()`

```{r}
head(log_skylight)
log_skylight <- skylight %>% mutate(log_intensity = log(intensity))
```

## Transforming your data

Create new column that gives you intensity in **Watts**!

Watts = (speed of light / wavelength in m) x photon count x Planck's constant

```{r}
c <- 3e+08 #speed of light
h <- 6.626e-34 #Planck's constant
nm_to_m <- 1e-09 #nm to m conversion
head(skylight)
```


## Transforming your data

```{r}
c <- 3e+08 #speed of light
h <- 6.626e-34 #Planck's constant
nm_to_m <- 1e-09 #nm to m conversion
head(skylight, 2)
skylight_W <- skylight %>% 
  mutate(intensity_W = (c/wavelength*nm_to_m)*(intensity)*h)
head(skylight_W, 2)
```


# Coding hack!

## Logical operators

$>$:  `>`

$<$:  `<`

$\geq$:  `>=`

$\leq$:  `<=` Not the same as `<-`!     

$=$: `==`

$\neq$: `!=`

and: `&`

or:  `|` Doesn't mean 'given that'!  

## Logical operators

```{r}
10 <= 6
```

## Logical operators

```{r}
10 <= 6
9 == 9
```

## Logical operators

```{r}
10 <= 6
9 == 9
9 != 7000000
```

## Logical operators

```{r}
10 <= 6
9 == 9
9 != 7000000
```

```{r eval = F}
9 = 8.999999
```

## Logical operators

```{r}
10 <= 6
9 == 9
9 != 7000000
```

```{r}
9 = 8.999999
```

## Logical operators

```{r eval = F}
10 < 11 & 10 > 9
```


## Logical operators

```{r}
10 < 11 & 10 > 9
```


## Logical operators

```{r}
10 < 11 & 10 > 9
```

```{r eval=F}
1 < 2 | 1 > 3
```


## Logical operators

```{r}
10 < 11 & 10 > 9
1 < 2 | 1 > 3
```


## Logical operators

```{r}
10 < 11 & 10 > 9
1 < 2 | 1 > 3
```

```{r eval = F}
10 < 11 & 10 < 9
```


## Logical operators

```{r}
10 < 11 & 10 > 9
1 < 2 | 1 > 3
10 < 11 & 10 < 9
```


## Logical operators

```{r}
10 < 11 & 10 > 9
1 < 2 | 1 > 3
10 < 11 & 10 < 9
```

```{r eval = F}
4 < 2 | 42 > 4
```


## Logical operators

```{r}
10 < 11 & 10 > 9
1 < 2 | 1 > 3
10 < 11 & 10 < 9
4 < 2 | 42 > 4
```


## Transforming your data with `filter()`

```{r}
str(skylight)
sky_twilight <- skylight %>% filter(sky_type == "twilight")
```


## Transforming your data with `filter()`

```{r}
head(sky_twilight,3)
sky_twilight <- skylight %>% filter(sky_type == "twilight")
```


## Transforming your data with `filter()`

```{r}
head(sky_twilight,3)
sky_twilight <- skylight %>% filter(sky_type == "twilight")
```


## Transforming your data with `filter()`

1. Filter the skylight dataset for 'natural' measurements -- anything that wasn't taken during exposure to urban skyglow.

2. BONUS CHALLENGE! Filter the dataset for skyglow and twilight measurements at wavelengths below 500 nm. 

(And here's a fun fact: these wavelengths cannot be detected by the satellites used to photograph the earth at night!)


## Transforming your data with `filter()`

1. Filter the skylight dataset for 'natural' measurements -- anything that wasn't taken during exposure to urban skyglow.

```{r}
sky_natural <- skylight %>% 
  filter(sky_type == "twilight" | sky_type == "starlight")
```

2. BONUS CHALLENGE! Filter the dataset for skyglow and twilight measurements at wavelengths below 500 nm. 

```{r}
sky_satellite <- skylight %>% 
  filter(wavelength < 500 & 
         (sky_type == "twilight" | sky_type == "skyglow"))
```


## Transforming your data with `filter()`

```{r}
ggplot(sky_satellite, aes(x = wavelength, y = intensity, 
                           color = sky_type)) + geom_line()
```

# Testing your test assumptions

## Testing for normality

```{r}
ggplot(chickwts, aes(x = weight)) + geom_histogram(bins = 7)
```


## Testing for normality
```{r eval = F}
shapiro.test(chickwts$weight) #you can do this

chick_weight <- chickwts$weight
shapiro.test(chick_weight) #or this, if you're fancy
```


## Testing for normality
```{r}
shapiro.test(chickwts$weight)
```


## Testing for normality
```{r echo=FALSE}
ggplot(chickwts, aes(x = weight)) + 
  geom_histogram(bins = 5) + facet_wrap("feed")
```

```{r eval=F}
ggplot(chickwts, aes(x = weight)) + 
  geom_histogram(bins = 5) + facet_wrap("feed")
```


## Testing for normality
```{r echo=FALSE}
ggplot(chickwts, aes(x = weight)) + 
  geom_histogram(bins = 5) + facet_wrap("feed")
```

Test animal (meatmeal, casein) and vegetable (everything else) feed types for normality separately!


## Testing for normality
```{r warning=FALSE}
chick_animal <- chickwts %>% 
  filter(feed == "meatmeal" | feed == "casein") %>%
  mutate(diet = "animal")
chick_veggie <- chickwts %>% 
  filter(feed != "meatmeal" & feed != "casein") %>%
  mutate(diet = "veggie")
shapiro.test(chick_animal$weight)
shapiro.test(chick_veggie$weight)
```


## Testing for normality
```{r warning=FALSE}
chick_animal <- chickwts %>% 
  filter(feed == "meatmeal" | feed == "casein") %>%
  mutate(diet = "animal")
chick_veggie <- chickwts %>% 
  filter(feed != "meatmeal" & feed != "casein") %>%
  mutate(diet = "veggie")
chick_diet = bind_rows(chick_animal, chick_veggie)
head(chick_diet)
```


## Testing for normality

```{r echo=FALSE}
ggplot(chick_diet, aes(x = weight)) + 
  geom_histogram(bins = 6) + facet_wrap("diet")
```


## Testing for equal variance
```{r eval=F, message=FALSE, warning=FALSE}
library(car)
leveneTest(weight~diet, data = chick_diet)
```

## Testing for for equal variance
```{r message=FALSE, warning=FALSE}
library(car)
leveneTest(weight~diet, data = chick_diet)
```

# Normal probability plots with `ggplot2`

## Normal probability plots with `ggplot2`
```{r}
chick_qq <- ggplot(chick_diet, aes(sample = weight)) + 
  geom_qq() + geom_qq_line()
```


## Normal probability plots with `ggplot2`
```{r}
chick_qq
```


## Normal probability plots with `ggplot2`
```{r}
chick_qq  + facet_wrap("diet")
```


## Normal probability plots with `ggplot2`
```{r}
chick_qq  + facet_wrap("diet") + coord_flip()
```


## Normal probability plots with `ggplot2`
```{r}
chick_qq_complete <- ggplot(chick_diet, aes(sample = weight)) + 
  geom_qq() + geom_qq_line() + facet_wrap("diet") + coord_flip()
```

# Non-parametric tests (part 1 of 2)

## Non-parametric tests (part 1 of 2)
Mann-Whitney U Test: the non-parametric equivalent of an independent two-sample t-test!

```{r}
# t.test(weight~diet, data = chick_diet)
wilcox.test(weight~diet, data = chick_diet) 
```

#Putting it all together

##Putting it all together

```{r}
titanic <- read.csv("titanic.csv")
head(titanic)
```

## Putting it all together

1. Test the fare of two groups (survivors and non-survivors) for normality
   a. Histogram (hint:`facet_wrap()`)
   b. Normal probability plot
   c. Shapiro-Wilk normality test
   d. Levene's test for equal variance
2. Make any data transformations necessary
3. Do a t-test on the transformed data
4. Compare the results with a Mann-Whitney U test


##Putting it all together

1a. Histogram

```{r fig.height=3.5}
ggplot(titanic, aes(x = fare)) + geom_histogram(bins=8) +
  facet_wrap("survived")
```


##Putting it all together

1b. Normal probability plot

```{r fig.height=3.5}
ggplot(titanic, aes(sample = fare)) + geom_qq() + geom_qq_line() +
  facet_wrap("survived") + coord_flip()
```

##Putting it all together

1c. Shapiro-Wilk normality test

First, split data by survival

```{r}
titanic_dead <- titanic %>% filter(survived == "died")
titanic_alive <- titanic %>% filter(survived == "survived")
```


##Putting it all together

1c. Shapiro-Wilk normality test

```{r}
shapiro.test(titanic_dead$fare)
shapiro.test(titanic_alive$fare)
```

##Putting it all together

1d. Levene's test for equal variance

```{r}
leveneTest(fare~survived, data = titanic)
```

##Putting it all together

2. Data transformations

```{r fig.height=2.5}
titanic_transformed <- titanic %>%
  mutate(fare_log = log(fare))
ggplot(titanic_transformed, aes(x = fare_log)) + geom_histogram(bins=8) +
  facet_wrap("survived")
```

##Putting it all together

2. Data transformations

Can't take the log of 0!

```{r}
titanic_transformed %>% filter(fare == 0) %>% head()
```


##Putting it all together

2. Data transformations

```{r fig.height=2.5}
titanic_transformed <- titanic %>%
  mutate(fare_log = log(fare + 1))
ggplot(titanic_transformed, aes(x = fare_log)) + geom_histogram(bins=8) +
  facet_wrap("survived")
```


##Putting it all together

3. Do a t-test on the transformed data

```{r}
t.test(fare_log~survived, data = titanic_transformed)
```

##Putting it all together

4. Compare the results with a Mann-Whitney U test

```{r}
# t.test(fare_log~survived, data = titanic_transformed)
wilcox.test(fare~survived, data = titanic)
```

# Homework time!
