---
title: "Contingency analysis for mtcars"
output: html_notebook
---
```{r}
library(tidyverse)
library(ggmosaic)
```

Is # of gears contingent on transmission type (automatic or manual?)

```{r}
str(mtcars)
```

`gear` is number of gears and `am` is transmission type.  According to the help file for `mtcars`, 0=automatic and 1=manual.

Both of these columns are numeric, so we need to convert them to factors in order to do any categorical data analysis.  We'll do that with `mutate()` and `as.factor()`

```{r}
mtcars2 <- #save as new object
  mtcars %>% 
  mutate(gear = as.factor(gear), #overwrite gear and am columns with factor versions
         am = as.factor(am)) %>% 
  mutate(am = forcats::fct_recode(am, manual = "1", automatic = "0")) # rename the levels of am so we don't have to remember what 0 and 1 mean

str(mtcars2)
```

After we've converted the columns to factors, we can make a contingency table

```{r}
mtcars.table <- table(transmission = mtcars2$am, gearnum = mtcars2$gear)
mtcars.table %>% addmargins()
```
It looks like automatic transmissions tend to have fewer gears!

Then we can do contingency analysis
```{r}
chisq.test(mtcars.table)
```
Looks like # of gears is contingent on transmission type


Then we can visualize the data with a mosaic plot
```{r}
ggplot(mtcars2) +
  geom_mosaic(aes(x = product(am, gear), fill = am)) +
  scale_fill_manual("Transmission Type:", values = c("blue4", "gold2")) + #changes the colors and legend name
  labs(x = "Number of Gears", y = "Transmission Type") + # adds labels
  theme_minimal() #adds a theme
```

