---
title: 'Week 11: Correlations'
author: "Avalon C.S. Owens, Eric R. Scott"
date: "11/16/2018"
output:
  ioslides_presentation:
    df_print: paged
    widescreen: yes
  pdf_document: default
---
```{r include=FALSE}
#required packages this week:
library(tidyverse)
```

## Outline

```{r message=FALSE, warning=FALSE, eval=FALSE}
library(ggplot2)
```

- Calculating correlation coefficient (Pearson and Spearman)
- Doing correlation test (Pearson and Spearman)
- Bi-variate scatter plots
- Anscombe's quartet
- Using `stat_summary()` in `ggplot2`




# Correlation Coefficients

## Correlation Coefficients

>- `cor()` takes two vectors and returns a correlation coefficient
>- Default is Pearson
>- Get Spearman correlation coefficient with `method = "spearman"`

```{r}
cor(trees$Girth, trees$Height)
cor(trees$Girth, trees$Height, method = "spearman")
```

## Correlation Coefficients

- Order of two vectors doesn't matter.

```{r}
cor(trees$Height, trees$Girth)
```

# Correlation Test

## Correlation Test

>- `cor.test()` takes two vectors and does a correlation test.
>- "`alternative = `" controls tails (`"two.sided"` is default)

```{r}
cor.test(trees$Height, trees$Girth)
```

## Correlation Test

>- `method = "spearman"` for Spearman correlation
>- Safe to ignore message: "Cannot compute exact p-value with ties"

```{r}
cor.test(trees$Height, trees$Girth, method = "spearman")
```

# Bi-variate Scatter Plots

## Scatterplots in ggplot2

>- Both `x` and `y` `aes`thetics need to be continuous
>- `geom_point()` plots points

```{r fig.height=3, fig.width=4}
ggplot(trees, aes(x = Height, y = Girth)) + geom_point()
```

## Scatterplots in ggplot2

>- OK to use either variable for the x-axis for correlations.
>- You can put `aes()` either inside of `ggplot()` or inside of a `geom`

```{r fig.height=3, fig.width=4}
ggplot(trees) + geom_point(aes(x = Girth, y = Height))
```

## Scatterplots in ggplot2

- Use `size = ` outside of `aes()` to change the size of all points

```{r fig.height=3, fig.width=4}
ggplot(trees) + geom_point(aes(x = Girth, y = Height), size = 4)
```

## Scatterplots in ggplot2

>- Put `size =` *inside* of `aes()` to map a variable to point size.
>- You probably won't have a use for this in this class.

```{r fig.height=3, fig.width=4}
ggplot(trees) + geom_point(aes(x = Girth, y = Height, size = Volume))
```

# Customizing plots

## Save previous plot

```{r}
p <- ggplot(trees) + geom_point(aes(x = Girth, y = Height, size = Volume))
```

## Finishing up your scatter plot

- Add units to axis labels
- (and to size legend)

```{r fig.height=3, fig.width=4}
p2 <- p +
  labs(x = "Girth (in)", y = "Height (ft)") +
  scale_size_continuous("Volume (ft^3)") 
p2
```

## Finishing up your scatter plot

- Try out a different `theme_*()`
- Type theme_ and browse the dropdown menu to try some out.

```{r fig.height=3, fig.width=4}
p2 + theme_classic()
```

# Anscombe's quartet

```{r echo=FALSE, fig.height=3, fig.width=10, message=FALSE, warning=FALSE}
anscombe %>% 
  add_column(ID = 1:11, .before = "x1") %>% 
  gather(-ID, key = "key", value = "value") %>% 
  separate(key, into = c("xy", "group"), sep = "(?<=\\w)") %>% 
  spread(key = xy, value = value) %>% arrange(group, ID) %>% 
  ggplot(aes(x = x, y = y)) + geom_point() + facet_wrap(~group, nrow = 1)
```

## Anscombe's Quartet

>- Anscombe's Quartet is a group of 4 bi-variate datasets
>- Similar mathematical properties despite very different shapes
>- Data visualization is important!
>- Available as `anscombe` in R
>- Even more mathematically similar and visually different datasets in the `datasauRus` package!

## Anscombe's Quartet

For your assigned set of x and y variables:

- Calculate the Pearson Spearman correlation coefficients
- Do a correlation test (just get the p-value, not all 6 steps)
- Make a bi-variate scatter plot
- Compare with another student

```{r rows.print=4}
anscombe
```

## `datasauRus`

```{r echo=FALSE, fig.height=5.5}
library(datasauRus)
datasaurus_dozen %>% 
  group_by(dataset) %>%
  mutate(dataset2 = paste(dataset, round(cor(x,y), 4))) %>% 
ggplot(aes(x = x, y = y, color = dataset)) +
  geom_point() +
  facet_wrap(~dataset2) +
  theme_minimal() + 
  theme(legend.position = "none")
```

## What now?

Work on homework?

*OR*

Go through a *way* quicker way to make bar plots and add significance letters?

```{r echo=FALSE, fig.height=4}
titanic <- read.csv("titanic.csv")
ggplot(titanic, aes(x = nationality, y = fare)) +
  stat_summary(geom = "bar", fun.y = mean) +
  stat_summary(geom = "errorbar", fun.data = mean_se, width = 0.25) +
  stat_summary(geom = "text", fun.y = mean, label = c("a", "b", "b"), size = 6,
               position = position_nudge(y = 6)) +
  annotate(geom = "text", x = 2, y = 40, label = "Only 4 lines of code!!!",
           angle = 335, size = 12, color = "red")
```


# Making complex plots simple with `stat_summary()`

## What's a `stat`?

- "geoms" add mapping to geometric features (e.g points, lines, bars)
- "stats" do some transformation.
- E.g. `geom_boxplot()` uses `stat = "boxplot"` by default.  Takes dataframe and calculates quartiles, fences, outliers.
- `geom_point()` uses `stat = "identity"` by default.  Simply passes x and y unchanged.

## Adding a stat

- Some geoms can take different stats, but usually not a good idea to change the default.
- You *can* add a `stat_*()` instead of adding a `geom_*()`
- E.g. instead of:

```{r eval=FALSE}
p + geom_boxplot(stat = "boxplot")
```

- You could do:

```{r eval=FALSE}
p + stat_boxplot(geom = "boxplot")
```

>- These are identical, no reason to *actually* do this.

## stat_summary() is flexible!{.smaller}

- Tell it how you want it to summarize your data frame and what geom you want it to use
- Remember bar plots?  First we summarized the data using `group_by(...) %>% summarize(...)`, then we plotted the summary data frame.
- With `stat_summary()` instead:

```{r fig.height=3}
ggplot(chickwts, aes(x = feed, y =  weight)) +
  stat_summary(geom = "bar", fun.y = "mean") + # makes the bars
  stat_summary(geom = "errorbar", fun.data = "mean_se", width = 0.25) # makes the errorbars
```

## stat_summary()

```{r, eval=FALSE}
ggplot(chickwts, aes(x = feed, y =  weight)) +
  stat_summary(geom = "bar", fun.y = "mean") # use x = feed, but y = mean(weight)
```

- `geom = "bar"` tells it to use `geom_bar()`
- `fun.y` is telling it how to get the y-values that `geom_bar()` needs.  Here, we used `mean()`
- Try changing `fun.y` to something else like `min`, `max`, or `median`.

## stat_summary()

```{r, eval=FALSE}
ggplot(chickwts, aes(x = feed, y =  weight)) +
  stat_summary(geom = "bar", fun.y = "mean") +
  stat_summary(geom = "errorbar", fun.data = "mean_se") 
# use x = feed, but ymin = mean(weight) - SEM, ymax = mean(weight) + SEM
```

- `geom = "errorbar"` tells it to use `geom_errorbar()`
- `mean_se` is a function from `ggplot2` that calculates ymin, and ymax aesthetics needed for errorbars.
- `fun.data` because `mean_se` returns a dataframe, not just the y-values

## Boxplots with letters using `stat_summary()`

- We can use `stat_summary()` to add letters to boxplots without making a new data frame

```{r}
titanic <- read.csv("titanic.csv") %>% filter(fare < 400)
p <- ggplot(titanic, aes(x = nationality, y = fare)) + geom_boxplot()
```
## Base boxplot

```{r}
p
```


## Adding letters with `stat_summary()`

```{r}
p + stat_summary(geom = "text", fun.y = "max", label = c("a", "b", "b"))
```

## Adjust letter position

- Nudge the letters up with `position = position_nudge()`

```{r}
p + stat_summary(geom = "text", fun.y = "max", label = c("a", "b", "b"),
                 position = position_nudge(y = 30)) #in units of y-axis
```

## Change the look of letters

- additional arguments get passed to `geom_text()`

```{r}
p + stat_summary(geom = "text", fun.y = "max", label = c("a", "b", "b"),
                 position = position_nudge(y = 30), #in units of y-axis
                 size = 5, color = "darkred") 
```

## Using `stat_summary()` in practice

- Making a summary table first, then plotting may be more *transparent*, easier for others to understand
- *But*, `stat_summary()` can make code cleaner, easier to read
- Complex plots with original data with multiple summary annotations (e.g. means, errorbars, significance letters) are made a **lot** simpler with `stat_summary()`
