---
title: "Recitation - Week 4"
output:
  word_document: default
  html_notebook: default
---

------ 1. Warm up ---------

Exercise 1. 

1) Download "Lizards.csv" data from Trunk.

2) Read in csv file to R

3) Calculate mean, standard deviation, variance, range and standard error of horn length of live vs killed horned lizards. 

4) Plot a histogram of horn length of live vs killed horned lizards. 

 - **Histogram binning**: Rules of Thumb (**always** use this guideline in this class!)

Sample Size       Number of Bins    
------------      --------------- 
Less than 16      Not Enough Data
16--31            5
32--63            6
64--127           7
128--255          8

```{r}

lizard<-read.csv("Lizards.csv")
head(lizard)
table(lizard$Survival)
anyNA(lizard$HornLength)

library(dplyr) # IMPORTANT!

summary.data<-lizard %>% filter(!is.na(HornLength)) %>% group_by(Survival) %>%       
                            summarise(average=mean(HornLength),
                                      sd = sd(HornLength),
                                      var = var(HornLength),
                                      range = max(HornLength) - min(HornLength),
                                      se = sd(HornLength)/sqrt(n()))
summary.data

bins <- function(x, bin.num){
  seq(min(x), max(x), (max(x) - min(x)) / bin.num)
}

living=subset(lizard, Survival=="living" & !is.na(HornLength))
killed=subset(lizard, Survival=="killed" & !is.na(HornLength))

par(mfrow=c(1,2))
hist(living$HornLength, breaks=bins(living$HornLength, 8),
     xlab="Horn Length (mm)", main="")
hist(killed$HornLength, breaks=bins(killed$HornLength, 5),
     xlab="Horn Length (mm)", main="")

```


----- 2. Introduce t.test function ----

*x* = for one sample t.test

*y* = optional (for two sample t.test)

*mu* = 0 for default (true mean for 1 sample, or difference if two sample) <- NULl HYPOTHESIS 

*alternative* = default is two.sided/non-directional (is the pop mean different from mu0), pop mean "greater" than sample, or pop mean "less" than sample mean. 

QUESTION: What if you were conducting a two sample t.test and you left the default where mu = 0 - what would your null hypothesis be? 
ANSWER: Null hypothesis is that your two samples have no differences in their means. 

*paired* = default is false (meaning samples are independent). TRUE means that each sample of each group is paired together, e.g. twin studies. 

*var.equal* = default is false, and therefore different variances for each group are calculated. Welsh


```{r}

bee<-read.csv("BeeNectar.csv")
head(bee)

# Example 1
bee.test<-with(bee, t.test(Pollen), mu=0) 
bee.test # explain all the output

names(bee.test) # outputs of model

bee.test$parameter
bee.test$estimate

bee.test$statistic # t statistic ... or the standard error of the mean, e.g. how well known is the mean based on the sample

# Example 2

lizard <- read.csv("Lizards.csv")
head(lizard)

with(lizard, t.test(HornLength ~ Survival))


```

----- 3. Introduce x-t functions ------

We can use other functions, qt and pt to ask questions about the sampling distribution for T tests. 

```{r}

# What is the probability density of critical values given degrees of freedom.... 

dt(0, df = 10) # probability T distribution function - returns the area under the T distribution 
pt(0, df = 10) # returns the cumulative area under the T distribution

# If we wanted to plot the probability density of critical values of the T distribution, given we know the degrees of freedom.
num<-seq(-3, 3, 0.1)
probs1<-dt(num, df=1)
probs2<-dt(num, df=5)
probs3<-dt(num, df=300)

plot(num, probs3, type="l", lty=1)
points(num, probs2, type="l", lty=2)
points(num, probs1, type="l", lty=3)

# What is the critical value, given we know the degrees of freedom and the significance level. This is the T table!
qt(0.85, df = 1) # T value in the 97.5% percentile is less than 1.96 if df = 524
qt(0.85, df = 1)


```


-------- Exercise 1. ---------

Problem: biologists were interested in whether worker size varies significantly among colonies across the landscape. 
  
1) If the population mean (mu) for bumble bee forager size (ITS) is 3.81, do you think your sample is representative of the whole population? In other words, are the population and sample means similar? If not, is your sample greater or less than the population mean?

2) Are the sample means for foraging worker size for colony 41 and colony 43 different from each other? If not, which colony has greater bumble bee worker size (report p-value)? Repeat question 2 for colony 32 and colony 2. 

3) Calculate the mean, standard deviation, standard error, and confidence intervals for forager size for each colony, except Colony 22. [Hint: dplyr might be good here]

4) Create three barplots for mean body size (ITS) for each colony (excluding colony 22) with error bars representing (i) standard deviation, (ii) standard error and (iii) 95% confidence intervals. [HINT: use arrows() function for creating error bars]

```{r}

bee<-read.csv("BeeNectar.csv")
head(bee)

m1<-with(bee, t.test(ITS, mu=3.81))
m1

m2<-with(bee, t.test(ITS, mu=3.81, alternative="greater")) # sample mean is greater than population mean. 
m2

m3<-with(bee, t.test(ITS, mu=3.81, alternative="less"))
m3

m4<-t.test(subset(bee, Colony=="41")$ITS, subset(bee, Colony=="43")$ITS)
m4

m5<-t.test(subset(bee, Colony=="32")$ITS, subset(bee, Colony=="2")$ITS, alternative="less")
m5

library(dplyr)

table(bee$Colony) # number of observations within each colony. Colony 22 does not have enough

# Solution 1
bee.m22<-subset(bee, Colony!="22")
bee.m22 %>% group_by(Colony) %>% summarise(mean=mean(ITS, na.rm=T),
                                       sd=sd(ITS, na.rm=T),
                                       se=sd(ITS, na.rm=T)/sqrt(n()),
                                       CI.upper=t.test(ITS)$conf.int[1],
                                       CI.lower=t.test(ITS)$conf.int[2])

#Solution 2
sums<-bee %>% filter(Colony!="22") %>% group_by(Colony) %>% summarise(mean=mean(ITS),
                                       sd=sd(ITS),
                                        se=sd(ITS)/sqrt(n()),
                                       CI.upper=t.test(ITS)$conf.int[1],
                                       CI.lower=t.test(ITS)$conf.int[2])
sums

par(mfrow=c(1,3))
barCenters<-with(sums, barplot(mean, names.arg=Colony, ylim=c(0,5)))
arrows(barCenters, sums$mean-sums$sd, barCenters, sums$mean+sums$sd, angle=90, code=3)

barCenters<-with(sums, barplot(mean, names.arg=Colony, ylim=c(0,5)))
arrows(barCenters, sums$mean-sums$se, barCenters, sums$mean+sums$se, angle=90, code=3)

barCenters<-with(sums, barplot(mean, names.arg=Colony, ylim=c(0,5)))
arrows(barCenters, sums$CI.upper, barCenters, sums$CI.lower, angle=90, code=3)


```

-------- Exercise 2. ---------

Problem: biologists were interested in whether introduced brook trout to affects the survival of chinook salmon. 

1) Does the presence or absence of introduced trout affect the proportion of chinook salmon surviving? 

2) Create three barplot for mean proportion surviving for both treatments with error bars representing (i) standard deviation, (ii) standard error and (iii) 95% confidence intervals. 


```{r}

salmon<-read.csv("BrookTrout.csv")
head(salmon)

salmon.t<-with(salmon, t.test(proportionSurvived ~ troutTreatment))
salmon.t

salmon.stats<-salmon %>% group_by(troutTreatment) %>% summarise(mean=mean(proportionSurvived),
                                                sd=sd(proportionSurvived),
                                                se=sd(proportionSurvived)/sqrt(n()),
                                                CI.upper=t.test(proportionSurvived)$conf.int[1],
                                                CI.lower=t.test(proportionSurvived)$conf.int[2])
salmon.stats

par(mfrow=c(1,3))
barCenters<-with(salmon.stats, barplot(mean, names.arg=troutTreatment, ylim=c(0,1), ylab="Proportion surviving"))
arrows(barCenters, salmon.stats$mean-salmon.stats$sd, barCenters, salmon.stats$mean+salmon.stats$sd, angle=90, code=3)

barCenters<-with(salmon.stats, barplot(mean, names.arg=troutTreatment, ylim=c(0,1),ylab="Proportion surviving"))
arrows(barCenters, salmon.stats$mean-salmon.stats$se, barCenters, salmon.stats$mean+salmon.stats$se, angle=90, code=3)

barCenters<-with(salmon.stats, barplot(mean, names.arg=troutTreatment, ylim=c(0,1), ylab="Proportion surviving"))
arrows(barCenters, salmon.stats$CI.upper, barCenters, salmon.stats$CI.lower, angle=90, code=3)

```



