---
title: "Recitation Week 1 Notes"
output:
  html_notebook: default
  pdf_document: default
---
## Welcome to Recitation (Natalie)
On powerpoint slides:

 - Format of class
 - Expectations
    - **collaborate** (help others, ask others for help)
    - be respectful to eachother
    - ask questions
    - tell us if we need to pause, slow down, repeat ourselves (we haven't been beginners for a long time)

## Orientation to RStudio (Natalie)

 - Console
 - Environment
 - Help pane

## Exercises to do directly in console (Natalie)
```{r}
#R does math
5+5
12/2
44*15
10^10
```
Introduce functions with `sqrt()` or `log()`
```{r}
sqrt(45)
```

Introduce assignment to objects
```{r}
x <- 5
5+x
x^x
sqrt(x)

x <- "biostats"
x

#vectors
x <- c(1, 2, 3, 4, 5, 25)
```

What happens if you do math to a vector?
```{r}
x * 3
```

Demonstrate how commented lines work
```{r}
x
#x
```

## Getting Help
Four ways:

 - `help(mean)`
 - `?mean`
 - search box in help pane
 - Google (stackexchange results especially)
```{r}
help(mean) # help("function") will retrieve the manual and help file for that function
?mean      # ?"function" will do the same as help()
```

## R Notebooks (Eric)
We'll go over the anatomy and use of an R notebook in recitation and the rest of the exercises will be done in code chunks of an R notebook.

Things to cover:
 - header and how to edit title and author names
 - text vs. code chunks
 - running code in a chunk (run button/shortcut or "play" button)
 - inserting a new chunk
 - knitting to Word doc

## Data Entry (Eric)
Before uploading data, you have to have all files in the same folder as this notebook for this code to run. 

```{r}
read.csv("SpiderAmputation.csv") # uploads your data
```
But if you want it to stick around, you need to assign to an object!

```{r}
data <- read.csv("SpiderAmputation.csv")
#or call it whatever you want
data
```

Always a good idea to check your data to make sure it loaded correctly
```{r}
head(data) # will call up the first 6 rows of your data frame
tail(data) # will call up the last 6 rows of your data frame
data[1:10, ] # data[rows, columns] therefore data[1:10, ] will call up rows 1 to 10, and all columns 
```

Other ways to check your data
```{r}
str(data) # shows what kind of data is in each column (character, integer, etc.). Helpful for spotting data entry errors
dim(data) # the dimensions of your data frame
nrow(data) # number of rows
ncol(data) # number of columns
length(data[ , 1]) # length of a vector or column in your data frame. You will have to specify the column if it is a data frame.
```
You can see the objects you've created in the Environment panel

## Descriptive Stats (Natalie)

```{r}
summary(data$speed)

mean(data$speed)
sd(data$speed)
var(data$speed)
min(data$speed)
max(data$speed)
range(data$speed) # IMPORTANT: in this class, "range" is a single value (min - max)

before<-subset(data, treatment=="before")
mean(before$speed)

after<-subset(data, treatment=="after")
mean(after$speed)
```

## Plotting (Eric)
Histograms
```{r}
#histograms only require specification of an x axis because the y axis is always frequency
hist(data$speed, xlab = "Speed", main = NULL)
```
Boxplots
```{r}
boxplot(data$speed, ylab = "Speed")
#If you want multiple boxplots side-by-side for different treatments, specify an x and y axis using a formula
boxplot(data$speed ~ data$treatment, xlab = "Treatment", ylab = "Speed")
```

## In-Class Exercise (Natalie)
 - on slides



## Common problems encountered in recitation


1. Remember to save your notebook in the same file that you saved your data. You may need to restart R Studio if you opened up the notebook in a different file

2. Thinking that your new subsetted dataframe is a vector

```{r}

data$after # this does not work since both "data" (containing both after and before treatment) and "after" (containing only after) are both dataframes. there is no "after" vector in "data"

# Instead you should do this:

data$speed
mean(data$speed)

after$speed
mean(after$speed)





```



