---
title: "Regression"
author: "Avalon C.S. Owens, Eric R. Scott"
date: "11/30/2018"
output:
  beamer_presentation: default
  ioslides_presentation:
    df_print: paged
    widescreen: yes
---


```{r include = F}
library(tidyverse)
library(ggthemes)
```

## Load in ur stuff!

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
lion <- read.csv("LionAge.csv")
head(lion)
```

## Plan for today

- Setting up a regression model with `lm()`
- Checking your regression assumptions
    - Residuals plots
    - Normal probability plots
- Doing a regression test with `summary()`
- Plotting best-fit lines with `geom_smooth()`

## A review of correlation

```{r}
ggplot(lion, aes(x = proportion.black, y = age)) + 
  geom_point() + theme_dark() + labs(x = "Proportion black on nose", y = "Age (yrs)")
```

## A review of correlation

```{r}
cor <- cor.test(lion$proportion.black, lion$age, method = "pearson")
cor$estimate #Pearson correlation coefficient
cor$estimate^2 #R-squared
```

## Setting up a regression model with `lm()`

$Y = \alpha + \beta*X$

```{r}
lion.model <- lm(age ~ proportion.black, data = lion) #lm stands for linear model
lion.model
```

## Setting up a regression model with `lm()`

Make your own regression model! 

See if you can predict tree girth as a function of tree height

```{r echo=F}
ggplot(trees, aes(x = Height, y = Girth)) + 
  geom_point() + theme_dark() + labs(x = "Tree height (ft)", y = "Tree diameter (in)")
```

## Setting up a regression model with `lm()`

Make your own regression model! 

```{r}
tree.model = lm(Girth ~ Height, data = trees)
tree.model
```

## Checking your regression assumptions (17.5)

1. LINEARITY: a true linear relationship between $X$ and $Y$ in the underlying population, such that $Y = \alpha + \beta*X$

2. NORMALITY: at every value of $X$, $Y$ is normally distributed

3. VARIANCE: at each value of $X$, the variance of $Y$ is equal

# Residuals plots

## Checking your regression assumptions

```{r echo=FALSE}
df <- data.frame(x = 1:20 + rnorm(20), y = 1:20 + rnorm(20))

df.fitted <-
  lm(y ~ x, data = df) %>%
  fortify() %>%
  mutate(ID = 1:20)

ggplot(df.fitted, aes(x = x , y = y)) +
  geom_point() +
  geom_line(aes(x = x, y = .fitted), color = "red") +
  geom_segment(aes(x = x, y = .fitted, xend = x, yend = .fitted + .resid))
```

## Checking your regression assumptions

```{r}
fortify(lion.model)
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(x = proportion.black, y = .resid)) + 
  geom_point()
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(x = proportion.black, y = .resid)) + 
  geom_point() + geom_hline(yintercept = 0)
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(x = proportion.black, y = .resid)) + 
  geom_point() + geom_hline(yintercept = 0, color = "red")
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(x = proportion.black, y = .resid)) + 
  geom_point() + geom_hline(yintercept = 0, color = "red") +
  labs(x = "Proportion black on nose", y = "Residuals")
```

## Checking your regression assumptions

Check the assumptions of your regression model with a plot of your own!

```{r echo=F}
ggplot(trees, aes(x = Height, y = Girth)) + 
  geom_point() + theme_dark() + labs(x = "Tree height (ft)", y = "Tree diameter (in)")
```

## Checking your regression assumptions

```{r}
ggplot(tree.model, aes(x = Height, y = .resid)) + 
  geom_point() + geom_hline(yintercept = 0, color = "red") +
  labs(x = "Height (ft)", y = "Residuals")
```

# Normal probability plots

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(sample = .resid)) + 
  geom_qq()
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(sample = .resid)) + 
  geom_qq() + geom_qq_line()
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(sample = .resid)) + 
  geom_qq() + geom_qq_line() +  coord_flip()
```

## Checking your regression assumptions

Check the assumptions of your regression model with a plot of your own!

```{r echo=F}
ggplot(trees, aes(x = Height, y = Girth)) + 
  geom_point() + theme_dark() + labs(x = "Tree height (ft)", y = "Tree diameter (in)")
```

## Checking your regression assumptions

```{r}
ggplot(tree.model, aes(sample = .resid)) + 
  geom_qq() + geom_qq_line(color = "darkgreen") + coord_flip()
```

## Checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(sample = .resid)) + 
  geom_qq() + geom_qq_line() + coord_flip()
```

## Transforming your data 

```{r}
lion.2 <- lion %>% mutate(age.2 = log(age))
lion.2
```

## Transforming your data 

```{r}
lion.king.2 <- lm(age.2 ~ proportion.black, data = lion.2) #lm stands for linear model
lion.king.2
```

## Re-checking your regression assumptions

```{r}
ggplot(fortify(lion.model), aes(sample = .resid)) + 
  geom_qq() + geom_qq_line() + coord_flip()
```

## Re-checking your regression assumptions

```{r}
ggplot(fortify(lion.king.2), aes(sample = .resid)) + 
  geom_qq() + geom_qq_line() + coord_flip()
```

## Doing a regression test with `summary()`

```{r}
summary(lion.model)
```

## Doing a regression test with `summary()`

```{r}
summary(lion.king.2)
```

# FYI...

## FYI...

You can also perform a regression test on a linear model with `anova()`

```{r paged.print=FALSE}
anova(lion.king.2)
```

## FYI...

You can calculate 95% confidence intervals for the slope using `confint()`

```{r}
confint(lion.king.2)
```

## Doing a regression test with `summary()`

Do a regression test of your own on your regression model!

Use the coefficients from the model to write the linear equation

```{r echo = F}
ggplot(trees, aes(x = Height, y = Girth)) + 
  geom_point() + theme_dark() + labs(x = "Tree height (ft)", y = "Tree diameter (in)")
```

## Doing a regression test with `summary()`

```{r}
summary(tree.model)
```


## Plotting best-fit lines with `geom_smooth()`

```{r}
ggplot(trees, aes(x = Height, y = Girth)) + 
  geom_point() + theme_dark() + labs(x = "Tree height (ft)", y = "Tree diameter (in)")
```

## Plotting best-fit lines with `geom_smooth()`

```{r}
ggplot(trees, aes(x = Height, y = Girth)) + geom_smooth(method = "lm") +
  geom_point() + theme_dark() + labs(x = "Tree height (ft)", y = "Tree diameter (in)")
```

## Plotting best-fit lines with `geom_smooth()`

```{r}
cool <- ggplot(trees, aes(x = Height, y = Girth)) + 
  geom_smooth(method = "lm", 
              level = 0.99,
              color = "dodgerblue", 
              fill = "pink") +
  geom_point(shape = 8, size = 2.5, color = "darkorange") + theme_economist() +
  labs(x = "Tree height (ft)", 
       y = "Tree diameter (in)", 
       title = "The coolest plot ever", 
       subtitle = "Aren't trees great?",
       caption = "Purple bands show 99% confidence interval")
```

## Plotting best-fit lines with `geom_smooth()`

```{r}
cool
```

# Changing figure sizes in R chunks

