---
title: "Power Analysis in R"
author: "Avalon C.S. Owens, Eric R. Scott"
date: "10/12/2018"
output:
  ioslides_presentation: default
  pdf_document: 
    latex_engine: xelatex
---

```{r include=FALSE}
library(emo)
library(pwr)
library(dplyr)
```

## Setup for today

- install the `pwr` package `r ji("package")`
- make a notebook and load it with `library(pwr)`
- download *oaks.csv* from Canvas and read it in with `read.csv()`

## Overview

- A note on citing packages `r ji("package")`
- Review 6-step hypothesis testing procedure
- Power analysis in R

# Citing R Packages

## Citing R Packages `r ji("package")`

- R packages are made by people!
- You can (and should) cite packages just like you would cite a scientific paper
- Find the citation with `citation()`

## Citing R Packages

"Power analyses were conducted with the *pwr* package in R (Champley 2018)."

```{r}
citation("pwr")
```


# Hypothesis testing

## Hypothesis testing

Whenever you are asked to test a hypothesis in this class, you **must** use the 6-step hypothesis testing procedure (unless we say otherwise).

1. Hypothesis

2. Choose and justify the test

3--5. Do the test! (in R)

6\. Summary statement

## Exercise 1

The city of Cambridge collects data on all city trees including location, diameter (in cm), and species.  These data are made available on [data.cambridgema.gov](https://data.cambridgema.gov/browse).  I downloaded the data and randomly sampled 20 oak trees growing in parks and 20 oak trees growing in cemeteries.  The data are available on Canvas.

**Test the hypothesis that oak trees grown in parks have a different diameter than oak trees grown in cemeteries.**



# Power Analysis

## `pwr` package

The `pwr` package has power analyses for many different tests:

- Two proportions/binomial, `pwr.2p.test()`
- T-tests, `pwr.t.test()`
- Chi-squared test, `pwr.chisq.test()`
- One way ANOVA, `pwr.anova.test()`
- Linear models, `pwr.f2.test()`
- Correlation, `pwr.r.test()`

## Power of a t-test

`pwr.t.test()` has 4 required arguments.  Define 3 of the 4 of them, and it will calculate and return the 4th

- `n` - sample size
- `d` - effect size
- `sig.level` - type I error probability (default 0.05)
- `power` - power of test

## Power of a t-test

Additional arguments define the type of t-test:

- `type` - one-sample, two-sample, or paired-sample t-test?
- `alternative` - same as in `t.test()`

## Power of a t-test

If we want to know the **power** of a test, what information do you need to supply to `pwr.t.test()`?

```{r, eval=FALSE}
pwr.t.test(n = ,
           d = ,
           sig.level = ,
           power = )
```

## Power of a t-test

```{r}
pwr.t.test(n = 15, d = 0.5, sig.level = 0.05) # power = ?
```


## Estimating required sample size

If you want to know the **sample size** required to detect a difference at some power, what do you need to supply to `pwr.t.test()`?

```{r, eval=FALSE}
pwr.t.test(n = ,
           d = ,
           sig.level = ,
           power = )
```


## Estimating required sample size

```{r}
pwr.t.test(d = 0.6, power = 0.8, sig.level = 0.05) # n = ?
```

## Power to detect differences in oaks

What is the power of the t-test we did to detect a difference between park and cemetery oaks?

- `n` = ?
- `sig.level` = ?
- `d` = ?

$$
d = \frac{\mid{\mu_1 - \mu_2}\mid}{\sigma}
$$
Estimate!

## Estimating parameters for `pwr.t.test`

```{r}
oaks <- read.csv("data/oaks.csv")
oaks %>% group_by(Location) %>% summarize(mean = mean(diameter), n = n())

d.est = (19.85 - 12.85)/sd(oaks$diameter)
d.est
```

- Is this a large or small effect size?

## Power analysis

```{r}
pwr.t.test(n = 20, d = d.est)
```

- Is this a lot of power?
- What is the probability we made a type II error?
- What could you do to increase your power?

## Learn more about power!

1. If you want to detect a medium effect size (d = 0.5) with power = 0.8 and a typical alpha of 0.05, what is the minimum sample size should you should have? (two-tailed, two sample t-test)



2. How big of an effect size (d) can you detect with a power of 0.6 and a sample size of 20 in a paired t-test?


3. Which has greater power, an independent two sample t-test or a paired t-test?

