---
title: "Week 7 Worked t-test"
output: html_notebook
---

# 1. Spooky Oaks

## Read in data

```{r}
oaks <- read.csv("data/oaks.csv")
str(oaks)
```

1. State hypotheses

H~0~: Park tree and cemetery tree diameters have the same population mean
H~A~: Park tree and cemetery tree diameters have different population means

H~0~: µ~1~ = µ~2~
H~A~: µ~1~ ≠ µ~2~


2. Choose Test
Two-sample two-tailed t-test. We are comparing two independent groups and looking for difference in either direction.

```{r}
oaks
t.test(diameter ~ Location, data = oaks)
```

3. t = 1.83
4. p = 0.0776
5. We fail to reject the null hypothesis

6. Mean oak tree diameters did not differ significantly between park and cemetery populations (two-sample two-tailed t-test, df = 30.474, t = 1.83, p = 0.0776).





## 2. Learn more about power!

1. If you want to detect a medium effect size (d = 0.5) with power = 0.8 and a typical alpha of 0.05, what is the minimum sample size should you should have? (two-tailed, two sample t-test)

```{r}
pwr.t.test(d = 0.5, power = 0.8)
```
**64**

2. How big of an effect size (d) can you detect with a power of 0.6 and a sample size of 20 in a paired t-test?

```{r}
pwr.t.test(n = 20, power = 0.6)
```
**0.71**

3. Which has greater power, an independent two sample t-test or a paired t-test?

```{r}
pwr.t.test(n = 30, d = 0.5)

pwr.t.test(n = 30, d = 0.5, type = "paired")
```
**Paired t-test**
