---
title: "Week 10 Recitation - Multiple comparisons"
output:
  html_notebook: default
  html_document: default
  pdf_document: default
  word_document: default
---

This week will be introducing the functions for multiple comparisons in R. We will introduce both Tukey's HSD and Bonferroni tests for ANOVA, as well as the Dunn's test for the Kruskal-Wallis test. 

```{r Packages }

# Old packages
library(ggplot2) # for ggplot()
library(ggformula) # stat_qqline() for creating the normality line in the QQ plot
library(dplyr) # for data wrangling
library(car) # leveneTest()

# New packages, so remember to install!
library(multcomp) # glht() for Bonferroni test
library(dunn.test) # dunn.test()

```

### Lecture 1. Circadian shift data

Here, we will be using the circadian shift data to learn multiple comparisons. This dataset compares phase shift in the circadian rhythm of melatonin production in participants given alternative light treatments (light exposed to eyes, knees, or the control). We want to know whether light treatment alters the circadian rhythm of melatonin production. 

```{r Enter data}

cir <- read.csv("CircadianShift.csv")

head(cir) # retrieves the first 6 rows of our data. This is preferred, than all your data. 
cir[1:10, ] # cir[rows, columns], so therefore I want to look at rows 1 to 10, and all columns (none specified means all)

boxplot(shift ~ treatment, data=cir) # let's look at the data 

```

### Lecture 2. Select appropriate model 

```{r Run models and test assumptions }

## We have three treatments. What test can we use?

cir.aov <- aov(shift ~ treatment, data = cir)

## Tests for the assumptions

# 1. Formal checks
shapiro.test(cir.aov$residuals) # normal data
leveneTest(shift ~ treatment, data=cir) # equal variances

# 2. Visual checks
cir %>% group_by(treatment) %>% summarise(mean = mean(shift),
                                          median = median(shift),
                                          var = var(shift)) 

shift.res <- data.frame(residuals = cir.aov$residuals)

ggplot(shift.res, aes(x = residuals)) + # histogram of residuals
  geom_histogram(bins = 5)

ggplot(shift.res, aes(sample = residuals)) +
  geom_qq() +
  stat_qqline() + # gives you the line to match against residuals to check for nomrality
  coord_flip() # swap axes (to match what Sara and the textbook provides you)

qqnorm(cir.aov$residuals, datax = TRUE); qqline(cir.aov$residuals, datax = TRUE) # non-ggplot way

## Do groups differ?

Anova(cir.aov) # gives you ANOVA table with F value, df and p-value (from car package)
anova(cir.aov) # (from base R)
#For this problem, Anova() and anova() give identical results, but for more complicated ANOVA models use Anova from the car package--it usually is more likely to give you the correct p-values.

```


### Lecture 3. Multiple comparisons

#### 3a. Tukey's test

Running all possible pairwise comparisons to delineate which groups are statistically different. 

```{r Running tukeys test and plotting}

TukeyHSD(cir.aov) # p adjusted tells you whether the two groups differ

boxplot(shift ~ treatment,
        data = cir,
        ylim = c(-3, 1),
        xlab = "Light treatment",
        ylab = "Circadian shift in melatonin production")

# Manually work out levels

letters <- c("a", "b", "a")

text(x = 1:3, y = 0.95, labels = letters)

```


** Reordering treatment in boxplot to make this more visually pleasing **

```{r reorder boxplot}

levels(cir$treatment) # this is the order
cir$treatment <- factor(cir$treatment, levels=c("control", "knee", "eyes")) # we can reorder them. HOWEVER - what if you misspell a word or use upper case "Control"? Spoiler: it replaces the factor with N/A values, so remember to spell and use upper/lower exactly the way the factors are used in the original data frame 

letters <- c("a", "a", "b")

boxplot(shift ~ treatment,
        data = cir,
        ylim = c(-3, 1),
        xlab = "Light treatment",
        ylab = "Circadian shift in melatonin production")

text(x = 1:3, y = 0.95, labels = letters, col="red")

## ggplot way:
ggplot(cir, aes(x = treatment, y = shift)) + #set up data and aesthetics
  geom_boxplot() + #add boxplot layer
  stat_summary(fun.y = "max", #what function to use to come up with the y values.  Try changing to "mean"
               geom = "text", #what kind of geom_ do we want to use? "label" is another alternative
               label = letters, #arguments passed to geom_text()
               position = position_nudge(y = 0.3))

  #stat_summary must be used instead of geom_text() because the latter acts on every single data point, not on the 3 points that we want.
```

#### 3b. Bonferroni test

This method allows for selecting a limited set of pairwise comparisons (the comparisons of interest) to reduce type 1 error.

```{r Bonferroni test}

shift.bonf <- glht(cir.aov, 
                   linfct = mcp(treatment = c("eyes - control = 0", # mcp is writing out your comparisons under your group name
                                              "knee - control = 0"))) # linear hypotheses (linfct) where your null hypotheses is eyes - control is 0

summary(shift.bonf, test = adjusted("bonferroni")) # bonferroni can change either the alpha or p-value. In lecture, we adjusted the probability of type 1 (alpha) error by dividing it by X number of comparisons. This test adjusts the p-value instead by multiplying it by X number of comaprisons. Both end up with the same result.

summary(shift.bonf, test = adjusted("none")) # p-values are not adjusted, therefore times 2 of the bonferroni adjusted values. 


```

### Exercise 1. Practicing multiple comparisons

*Data*: The counts of insects in agricultural experimental units treated with different insecticides (sprays = A, B and C). Let's say that C spray was an organic insectide (like tobacco spray) and the other two are synthetic chemicals.

*Instructions:*
1. Run an ANOVA (assume that all assumptions have been met).
2. Conduct a multiple comparison that you think is relevant. 
3. Plot your results.
4. Write a summary statement for both the ANOVA and multiple comparison tests. 

```{r Exercise 1}

insect <- read.csv("InsectSpray.csv")
head(insect)

# 1. Run ANOVA
spray.aov <- aov(count ~ spray, data = insect)
anova(spray.aov)

# 2. Multiple comparisons
TukeyHSD(spray.aov)

spray.bonf <- glht(spray.aov, # if we wanted to compare organic C spray with the other two 
                   linfct = mcp(spray = c("A - C = 0",
                                          "B - C = 0")))
summary(spray.bonf, test = adjusted("bonferroni"))

# 3. Plot your results
boxplot(count ~ spray, 
        ylim = c(0, 30),
        xlab="Spray", 
        ylab = "Insect count", 
        data = insect)


letters <- c("a", "a", "b") # results from Tukey

text(x = 1:3, y = c(25, 23, 10), labels = letters, col="blue")



```
** 4. Write a summary statement **

The mean insect counts were not equal across the different insect sprays (one-way ANOVA, F = 44.636, df = 2, P < 0.001). The C spray had significantly less insect counts than the other two sprays (Tukey's HSD, all p < 0.0001), yet the mean insect counts for A and B sprays were not statistically different (Tukey's test, P = 0.86).


### Lecture 4. Non-parametric multiple comparisons

```{r Dunns test}

cir.kruskal <- kruskal.test(shift ~ treatment, data = cir)
cir.kruskal

dunn.shift <- with(cir, dunn.test(x = shift,  # x is response
                    g = treatment,  # g is "group"
                    altp = TRUE,  # FALSE if we want alpha = alpha/2
                    list = TRUE,   # provides us with another long list similar to the other multiple comparison tests
                    method = "bonferroni")) # many family-wise error rates (FWER) options for choosing where this test lies on the type 1/type 2 seesaw

names(dunn.shift) # provides you with the names of the outputs of the function, so that you can use these to extract the outputs you want.

dunn.shift$Z # Z test statistic
dunn.shift$P.adjusted # P-values

```


### Exercise 2. 

*Data*: Daily air quality measurements on Roosevelt Island, New York in 1973. Test whether the mean ozone (ppb) was different among the three months. 

*Instructions:* 
1. Run a Kruskal-Wallis test
2. Conduct a Dunn's test 
3. Plot your results
4. Write a summary statement for both the Kruskal-Wallis and Dunn's tests. 

```{r Exercise 2}

quality <- read.csv("AirQuality.csv")

# 1. Kruskal-Wallis
kruskal.test(Ozone ~ as.factor(Month), data = quality)

# 2. Dunn's test
with(quality, dunn.test(x = Ozone, g = as.factor(Month),  # x is response and g is "group"
                    altp = TRUE, # we want the alpha to be 0.05
                    method = "bonferroni",  
                    list = TRUE)) 

# 3. Plotting your results
letters <- c("a", "b", "a") # same results as ANOVA

boxplot(Ozone ~ as.factor(Month), 
        ylim = c(0, 200),
        xlab="Month", 
        ylab = "Ozone", 
        xaxt = "n",
        data = quality)

axis(1, at=1:3, labels = c("May", "July", "September"))
text(x = 1:3, y = 150, labels = letters, col="blue")



```

**4. Write a summary statement**

The three months in 1978 did not have the same ozone concentrations (ppb) (Kruskal-Wallis test, $\chi^{2}$ = 20.705, df = 2, P < 0.001). July had higher ozone concentrentrations than all other months (Dunn's test, all p < 0.005), yet May and September did not differ in their ozone concentrations (Dunn's test, P = 0.518).

