---
title: "Sampling and Summarizing Data"
author: "Eric R. Scott, Avalon Owens"
output:
  ioslides_presentation:
  pdf_document: default
---
```{r}
library(tidyverse)
```

## Overview
- probability distributions (uniform, binomial)
- plotting probability distributions (`geom_density()`, `geom_vline()` for adding marks for p > 0.05 and stuff)
- standard error (write a function for this?)
- confidence intervals (calculated using `dt()`---or is it `pt()`?.  Anyway, don't introduce `t.test()` yet)
- `geom_bar()`(which does calculations for you), `geom_col()` (if you have means already calculated), `geom_errorbar(stat = mean_se)`.



easy way to make errorbars:
```{r}
ggplot(mtcars, aes(cyl, qsec)) + 
    stat_summary(fun.y = mean, geom = "bar") + 
    stat_summary(fun.data = mean_se, geom = "errorbar")
```

more transparent way to make errorbars:
```{r}
# make a function to calculate standard error
sem <- function(x){
  sd(x)/sqrt(length(x))
}

plotdata <- mtcars %>% 
  group_by(cyl) %>% 
  summarize(meanq = mean(qsec),
            errmin = meanq - sem(qsec),
            errmax = meanq + sem(qsec))

ggplot(plotdata, aes(x = cyl, y = meanq)) +
  geom_col() +
  geom_errorbar(aes(ymin = errmin, ymax = errmax))
```


## Warmup
 - Take a random sample of 15  *integers* from a sequence of *integers* ranging from 1 to 1000.
 - Take a random sample of 15 numbers from a normal distribution with a mean of 500 and a standard deviation of 10
 - What are some of the differences between sampling numberse these two ways?

### Discrete probability distributions
Remember to plot these as barplots with space between the bars, not as histograms or as lines

#### Uniform
`dunif()` gives the probability of a single event, e.g. probability of rolling a 6 on a single die. This is pretty easy to do with math instead.
```{r}
#die
dunif(5, min = 0, max = 6)
#Naming arguments can be helpful, but you don't have to do it if you supply them in the same order as in the helpfile
dunif(5, 0, 6)
1/6
```

`punif()` gives the probability distribution function. E.g. P[X < 5] or P[X > 3]. 

- Read the help file to figure out what `lower.tail` does.
```{r}
#proability of rolling 1 through 4
punif(4, min = 0, max = 6, lower.tail = TRUE)

#probability of rolling greater than 4
punif(4, min = 0, max = 6, lower.tail = FALSE)
```
All of this is easy to do with math, but all probability distributions use similar functions.  That is, there's a `pbinom()`, a `pnorm()`, a `ppois()` (for poisson, which we won't cover), etc.

#### Plotting a discrete probability distribution
How does `barplot()` work?
Look at help file. Notice that there are *MANY* arguments, but that all but one have defaults.  That means you don't need to worry about them!  Notice that rather than x and y it wants "height".  That means it isn't going to calculate anything for you the way `boxplot()` does!
```{r}
#make a vector of heights
(probs <- dunif(1:6, min = 0, max = 6))  #parentheses make it show the result AND assign to object
barplot(height = probs, 
        names.arg = 1:6,         #add names.arg to get x-axis labels
        ylim = c(0,1),           #xlim and ylim change the range of the axes.  
        ylab = "Probability",
        xlab = "number of pips on a die")
```

#### Binomial Distribution
Number of times one of two mutually exclusive events occurs in n independent trials
Take a look at the helpfile for `dbinom()`

 - Which argument is #trials? (size)
 - which argument is binomial p AKA p[success] ? (prob)

10 coin tosses, P[X = 5 heads] 
```{r}
dbinom(x = 5, size = 10, prob = 0.5)

```

probability of germinating titan arum seeds
```{r}
p <- 0.30
n <- 3
x <- 0:3

(probs <- dbinom(0:3, n, p))

barplot(height = probs, names.arg = 0:3,
        xlab = "# sucessful germinations",
        ylab = "probability",
        ylim = c(0,1))
```

Play with shape of binomial distribution
```{r}
p = 0.3
n = 40

x = 0:n
probs <- dbinom(x, n, p)
barplot(probs, names.arg = x, ylim = c(0,1),
        ylab = "Probability",
        xlab = "# successes")
```

**Advanced** barplots. Recreating a figure from lecture slides showing how binomial distributions change when parameters change.
```{r}
n = 20
x = 0:n
p1 <- dbinom(x, n, 0.1)
p2 <- dbinom(x, n, 0.5)
p3 <- dbinom(x, n, 0.9)

#add = argument allows overplotting.  Not perfect because green bars cover red bars and blue covers green not side-by side.
#gold star for anyone who figures out how to do this correctly!
barplot(p1, col = "red", ylim = c(0, 0.3), names.arg = 0:n)
barplot(p2, col = "green", add = TRUE, axes = FALSE)
barplot(p3, col = "blue", add = TRUE, axes = FALSE)

###ggplot2 way.  To run this, you might need to install ggplot2 using install.packages("ggplot2")


df <- data.frame(probs = c(p1, p2, p3), #one column is all the calculated probabilities
                 ps = rep(c("0.1", "0.5", "0.9"), each = 21), #one column is the probabilities of success
                 xs = rep(0:20, 3)) #one column is just 0 - number of trials to be used as the x-axis

library(ggplot2)
ggplot(df) +
  geom_col(aes(y = probs, x = xs, fill = ps), position = "dodge") +
  #position = "dodge" makes the bars not overlap,
  #fill = makes it color code by that variable
  xlab("# of successes") +
  ylab("Probability")
```

#### Exercise 1

 - What's the probability of drawing an ace of spades from a deck of cards (with no Jokers)?
 - What's the probability of drawing 5 aces (of any suit) in a row from a deck of cards (with replacement!)?
 
```{r}
#prob of 5 aces out of 5 draws (with replacement)

n = 5
p = 4/52
x = 5
dbinom(x, n, p)
```

### Continuous distributions
probability is area under the curve

#### Normal

- what are the defaults for mean and sigma for `pnorm()`? which normal distribution is this?
- Don't use `dnorm()`!
```{r}
#P[X = x] is always 0
dnorm(1)
#doesn't work like expected!  Don't use dnorm!
```
- what % of curve is between -2 and 2 standard deviations?
```{r}
pnorm(2) #only lower tail, not between. Draw on board
pnorm(2) - pnorm(-2) #subtract the lower bit
```

#### Plotting continuous distributions
You could use points to approximate the shape
```{r}
xvals <- seq(-5, 5, 0.2) #just a sequence of x values to get probabilities for
plot(x = xvals, y = dnorm(xvals),
     ylab = "Probability Density",
     xlab = "Z",
     type = "l") #line plot instead of points
```

Another way is to use `curve()` which takes a function as the first argument
```{r}
curve(dnorm(x), from = -5, to = 5,
      ylab = "Probability Density",
      xlab = "Z")
#could also use add = TRUE to overlay ontop of scatter plot
```
**Advanced** plotting: adding vertical lines and colors and arrows and text to a plot
```{r}
curve(dnorm(x), from = -5, to = 5,
      ylab = "Probability Density",
      xlab = "Z",
      col = "blue")
abline(v = -2, col = "red")
abline(v = 2, col = "red")
text(x = 0, y = 0.05, labels = "95%")
arrows(x0 = c(-0.5, 0.5), y0 = c(0.05, 0.05), x1 = c(-2, 2), y1 = c(0.05, 0.05))
```

### Confidence intervals and SEM

#### SEM
 - sd describes spread in data in same units as mean
 - SEM is a new summary statistic that describes the precision of your estimate of the mean

$$SEM = \frac{s}{\sqrt{n}}$$

We're going to use a built-in data set calld `ToothGrowth`.  There are many built-in datasets which you can view with the command `library(help = "datasets")`
```{r}
ToothGrowth
str(ToothGrowth)
```


There is no built-in SEM function in R.  You can calculate it though, or you can **write your own function!**
```{r}
(mu <- mean(ToothGrowth$len))
(s <- sd(ToothGrowth$len))
(n <- length(ToothGrowth$len))
(sem <- s / sqrt(n)) 

SEM <- function(x){
  sd(x)/sqrt(length(x))
}
```

#### Confidence intervals
Confidence intervals give an upper and lower confidence limit at some level of $\alpha$, usually 95%.  A 95% confidence interval means if you were to sample from a population repeatedly, 95% of your sample means would fall into those limits. R calculates this using the `t.test()` function

*approximate*
```{r}
conf.lower<-mu + 1.96*SEM(ToothGrowth$len)
conf.upper<-mu - 1.96*SEM(ToothGrowth$len)
```

*exact*
```{r}
t.95 <- t.test(ToothGrowth$len)

#objects created by t.test have a bunch of things in them.  You can see what's there with names()
names(t.95)
#or with autocomplete after typing t.95$

as.numeric(round(t.95$conf.int, 1))

upper <- t.95$conf.int[1]
lower <- t.95$conf.int[2]
```
#### Exercise 2

 - get means and SEMs for len of each dose using `subset()` or `group_by()` and `summarise()` from the `tidyverse` (more advanced)

```{r}
#probably not a good idea to overwrite built-in data, but I'm doing it here for simplicity
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

low <- subset(ToothGrowth, dose == 0.5)
mid <- subset(ToothGrowth, dose == 1)
high <- subset(ToothGrowth, dose == 2)

means <- c(mean(low$len), mean(mid$len), mean(high$len))
SEM <- function(x){
  sd(x)/sqrt(length(x))
}
SEMs <- c(SEM(low$len), SEM(mid$len), SEM(high$len))
means
SEMs
```


**Tidyverse way**
See more about how this works using the dplyr cheatsheet in the help menu > cheatsheets > Data Manipulation with dplyr, tidyr
```{r}
library(tidyverse)
Tooth.Summary <- ToothGrowth %>% 
  group_by(dose) %>% 
  summarise(mean.len = mean(len), SEM.len = sd(len)/sqrt(length(len)))
Tooth.Summary
```

#### Plotting error bars
Barplots generally represent means as the heights of the bars so we need to summarise data before making a bar plot.
```{r}
barplot(means, names.arg = levels(ToothGrowth$dose),
        ylab = "odontoblast length (nm)",
        xlab = "vitamin C dose (mg/day)")

###OR###
# barplot(Tooth.Summary$mean.len, names.arg = Tooth.Summary$dose,
#         ylab = "odontoblast length (nm)",
#         xlab = "vitamin C dose (mg/day)")
```
#### Adding error bars
`barplot()` has a weird feature that when you assign it to an object, the object contains values for the centerpoints of the bars, which is useful for plotting

```{r}
barCenters <- barplot(means, names.arg = levels(ToothGrowth$dose),
        ylab = "odontoblast length (nm)",
        xlab = "vitamin C dose (mg/day)",
        ylim = c(0, max(means) + max(SEMs)))
barCenters
arrows(barCenters, means - SEMs, barCenters, means + SEMs, angle = 90, code = 3)

```

#### Exercise 3
 - make a bar plot like this but with error bars representing 95% confidence intervals instead of SEMs

```{r}
low.ci <- as.numeric(t.test(low$len)$conf.int)
mid.ci <- as.numeric(t.test(mid$len)$conf.int)
high.ci <- as.numeric(t.test(high$len)$conf.int)

CIs <- as.data.frame(rbind(low.ci, mid.ci, high.ci))

barCenters <- barplot(means, names.arg = levels(ToothGrowth$dose),
        ylab = "odontoblast length (nm)",
        xlab = "vitamin C dose (mg/day)",
        ylim = c(0, max(CIs)))
barCenters
arrows(barCenters, CIs$V1, barCenters, CIs$V2, angle = 90, code = 3)
```

