---
title: "Probability Distributions and Bar Plots"
author: "Eric R. Scott, Avalon C.S. Owens"
header-includes:
  - \usepackage{coloremoji}
output:
  pdf_document: default
  ioslides_presentation: default
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(cowplot)
library(ggplot2)
library(dplyr)
library(knitr)
knitr::opts_chunk$set(error = TRUE)
```

## Overview

- Debugging common errors
- Beautifying your homework
- Probability distribution functions in R
- Plotting binomial distributions with `ggplot2`
- Building on the boxplot: more fun with `ggplot2`

## My homework won't knit! 😖

- When you click "Knit", R instantly forgets all stored datasets and closes all opened packages
- R starts from a clean slate and reads your code top to bottom
- Your code has to run with ZERO ERRORS in order for it to knit!
- Check for errors with "Run" → "Restart R and run all chunks"
- If there is an error, the error message will be written in <span style="color:red">red</span>

## I can't read in the file! 😭

- You must save your .Rmd document for it to exist somewhere
- Your .csv file needs to be in the same place as the .Rmd (to read it in with only the file name)
- Simple typos in the file name will give you an error like this:

```{r}
read.csv("Camridge Trees.csv")
```

## Tips on data management

- Create a folder for Biostats recitation
- Create a separate sub-folder for each week
    + Make a new R Notebook and save it in this folder
    + Save data (.csv files) you need in the same folder

# Beautifying your Homework

## What we're doing to help

- Improved formatting in PS3.Rmd
- Questions as headers, with supporting text in *italics*
- Minimal text---refer to the handout .pdf for more information
- First code chunk that loads packages and sets some options (**edit at your own risk**)

## Tips on code beautification

- Start new lines at natural break points (e.g. after pipes)
- Use indents (you can also "auto-indent" with command + i)

**UGLY:**

```{r eval=FALSE}
chickwts %>% group_by(feed) %>% summarize(mean = mean(weight), sd = sd(weight), range = max(weight) - min(weight))
```

**BEAUTIFUL:**

```{r eval=FALSE}
chickwts %>%
  group_by(feed) %>%
  summarize(mean = mean(weight),
            sd = sd(weight),
            range = max(weight) - min(weight))
```

## Tips on text formatting

- After knitting to Word, make sure to check the formatting; directly edit the .docx to improve readability, if you need to
- After printing, circle your answers if they're hard to find
- Check your rounding! If `round()` isn't working, you can always just write out your answer below in plain text

```{r}
weights <- c(2,2,3)
mean(weights)
```

**mean weight = 2.3 g**

# Probability distributions

```{r echo=FALSE, fig.height=3}
unif.df <- data.frame(x = 1:6, prob = dunif(1:6, min = 0, max = 6))
unif.plot <- ggplot(unif.df, aes(x = x, y = prob)) +
  geom_col() +
  ylim(0,1) +
  labs(y = "P[x]",
       x = "Number of pips on a die",
       title = "Uniform") +
  theme_grey()

binom.df <- data.frame(x = 1:20, prob = dbinom(1:20, size = 20, prob = 0.5))
binom.plot <- ggplot(binom.df, aes(x = x, y = prob)) +
  geom_col() +
  ylim(0,1) +
  labs(y = "P[x]",
       x = '# of "heads" in 20 coin flips',
       title = "Binomial") +
  theme_grey()

norm.plot <- ggplot(data.frame(x = c(-10, 10)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 2)) +
  labs(y = "P[x]",
       x = "x",
       title = "Normal") +
  theme_grey()

plot_grid(unif.plot, binom.plot, norm.plot, nrow = 1, ncol = 3)
```

## Probability Distributions in R

- All probability distributions in R have four related functions:
    + A density function that starts with `d` (e.g. `dbinom()`)
    + A distribution function that starts with `p` (e.g. `pbinom()`)
    + A quantile function that starts with a `q` (e.g. `qbinom()`)
    + A random number function that starts with `r` (e.g. `rbinom()`)

- All distributions work similarly
- Today we'll focus on binomial and normal distributions
- Next week we'll introduce the t distribution!

## Binomial Distribution

Number of times one of two mutually exclusive events occurs in n independent trials

```{r echo=FALSE, fig.height=3.5, fig.width=7}
arums <- data.frame(seeds = 0:3)
arums <- arums %>% 
  mutate(prob = dbinom(seeds, size = 3, prob = 0.30))
#using mutate to get the probability column instead

arumplot <- ggplot(arums, aes(x = seeds, y = prob)) +
  geom_col() +
  ylim(0,1) +
  labs(x = "# of successful germinations", y = "P[x]") +
  theme_gray() +
  ggtitle("Probability of Successful Germination of Titan Arum Seeds")
arumplot
```

## `*binom()` family of functions

- `dbinom()`, the density function---what's the probability of $[X = 5]$?
- `pbinom()`, the distribution function---what's the probability of $[X \leq 5]$?
- `qbinom()`, the quantile function---what quantile is associated with $P = 0.62$?
- `rbinom()`, the random function---what's a random number from a binomial distribution with $n = 10$ and $p = 0.5$?

## `*binom()` Think-Pair-Share
- Take a look at the help file for `pbinom()`

    + Which argument is the number of trials?
    + Which argument is binomial p AKA p[success]?
    + What does `lower.tail=FALSE` or `lower.tail=F` do?
    + Which `*binom()` function would you use to figure out the probability of tossing a coin 10 times and getting 5 heads?

## `*binom()` Arguments

- `size =` is $n$
- `prob =` is the binomial probability, $p$
- `lower.tail = ` switches between $P[X ≤ x]$ (`TRUE`, default) and $P[X > x]$ (`FALSE`)
- `dbinom()`, the density function, would be used to find the probability of getting 5 heads in 10 coin tosses $P[X = 5]$

```{r}
dbinom(x = 5, size = 10, prob = 0.5) %>% round(4)
```

## `pbinom()`, the distribution function

- If you toss a coin 10 times, what's the probability of getting 5 or fewer heads?
- $P[X \leq 5$ heads$]$

```{r}
pbinom(q = 5, size = 10, prob = 0.5) %>% round(4)
```

- more than 5 heads?
- $P[X > 5$ heads$]$

```{r}
pbinom(q = 5, size = 10, prob = 0.5, lower.tail = F) %>% round(4)
```

## Practice binomial calculations

1. If you roll a single 6-sided die 10 times, what's the probability of rolling a 6 more than once?

2. A study of loggerhead turtles estimates the annual survival probability of juveniles at 0.73. With a starting population of 1000 juvenile turtles, what's the probability that at least 700 will survive?

```{r eval=FALSE, include=FALSE}
pbinom(699, 1000, 0.73, lower.tail = FALSE)
```


## Normal Distribution

Defined by mean and standard deviation

```{r echo=FALSE}
norm.plot +
  stat_function(fun = dnorm, args = list(mean = 2, sd = 1), color = "red") +
  stat_function(fun = dnorm, args = list(mean = -3, sd = 3), color = "green") +
  ggtitle("A Few Normal Distributions")
```

## `*norm()` family of functions

- ~~`dnorm()`~~, the density function---doesn't do what you might think it does, so **don't use it!**
- `pnorm()`, the distribution function---what's the probability $[X > 1.64]$?
- `qnorm()`, the quantile function---what quantile is associated with $P = 0.05$?
- `rnorm()`, the random function---what is a random number from a normal distribution with $\mu = 0$ and $\sigma^2 = 1$?

## `pnorm()`, the distribution function

Lower tail probability is the default.

```{r}
pnorm(1.96, mean = 0, sd = 1)
```

This is the area under the curve **less** than 1.96

To get the upper tail:

```{r results="hold"}
1 - pnorm(1.96)
#or
pnorm(1.96, lower.tail = FALSE)
```

## Practice normal calculations

1. $P[X > 1.2]$ with $\mu = 1, \sigma^2 = 0.5$

2. $P[Z > 1.2]$

3. $P[Z \geq 1.2]$

## Uniform Distributions

- ...Are boring! 😴
- Probabilities are easy to calculate by hand or with a calculator
- However, they can be useful for generating random numbers

## Random numbers from uniform

You put a 100 m long transect through your field site and you want to choose a three random locations along the transect to put your experimental plots.

- Why would `rnorm()` be a bad choice?

```{r}
runif(3, min = 0, max = 100)
```

- `sample()` is probably a better choice for random integers

# Coding hack #1

## Easy sequences

```{r}
1:10
```
```{r}
sample(1:1000, 3)
```

# Bar plots with `ggplot2`

## Plotting a binomial distribution

Plot the binomial distribution that represents the possible outcomes of 6 coin flips

- Because binomial distributions are **discrete**, we use bar plots
- To plot a binomial distribution, first we will need a data frame!

## Making a data frame

- We can create a data frame from scratch with `data.frame(column_name = data)`
- All of the columns have to have the same length!

```{r}
coinflips <- data.frame(num.heads = 0:6,
                        probs = dbinom(0:6, size = 6, prob = 0.5))
coinflips
```

## Making a bar plot

- use `geom_col()` when you've already know the heights of the bars you want to plot
- use `geom_bar()` if you want ggplot to calculate group means and use them for bar heights

Since we want to plot the probabilities as the heights, we will use `geom_col()`. (We might practice using `geom_bar()` next week.)

## Fill in the blanks

```{r, eval=FALSE}
ggplot(???, aes(x = ???, y = ???)) +
  geom_col()
```

## Fill in the blanks

```{r}
ggplot(coinflips, aes(x = num.heads, y = probs)) +
  geom_col()
```

## Finishing touches

- Label your axes!
- Since probability ranges from 0--1, we might want to adjust the scale of the y-axis using this coding hack: `ylim(0,1)`

```{r}
p <- ggplot(coinflips, aes(x = num.heads, y = probs)) +
  geom_col() +
  labs(x = "Number of heads in 10 coin flips", y = "Probability") +
  ylim(0,1)
```

## Finished!

```{r}
p
```

# Building on the boxplot

```{r echo=FALSE, fig.height=3, fig.width=7}
ggplot(chickwts, aes(x = feed, y = weight)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.color = "red") +
  # geom_jitter(alpha = 0.3, width = 0.1) +
  theme_grey() +
  labs(x = "Feed", y = "Weight (g)")
```

## Multiple `geom`s

- Last week we introduced `geom_violin()` and `geom_jitter()`
- Sometimes boxplots can be improved by adding additional representations of data
- Boxplot + jitter plot shows boxplot info plus the actual data
- Boxplot + violin plot is nice when there are many data points

```{r}
p2 <- ggplot(chickwts, aes(x = feed, y = weight)) +
  labs(x = "Feed", y = "Weight (g)")
```

## Boxplot + jitter

- Nice when the number of points is small

```{r}
p2 + geom_boxplot() + geom_jitter()
```

## Making adjustments

- make points transparent with `alpha = `
- make jitter narrower with `width = `
- change boxplot outlier points so they look different than data

```{r fig.height=3}
p2 + geom_boxplot(outlier.color = "red", outlier.shape = "triangle") + 
  geom_jitter(width = 0.25, alpha = 0.3)
```

## Order of `geom` matters

```{r fig.height=2, fig.width=4}
p2 + geom_jitter() + geom_boxplot()
p2 + geom_boxplot() + geom_jitter()
```

## Boxplot + violin

- Adjust boxplot width to fit inside violins
- Use `trim = FALSE` in `geom_violin()` to prevent it from truncating the ends

```{r fig.height=3}
p2 + geom_violin(trim = FALSE) +
     geom_boxplot(width = 0.1)
```



