---
title: "Multiple comparisons in R"
author: "Avalon C.S. Owens, Eric R. Scott"
date: "11/09/2018"
output:
  ioslides_presentation:
    widescreen: yes
  pdf_document: default
  html_document:
    df_print: paged
  beamer_presentation: default
---
Note: This was edited *after* the recitation to do Bonferroni tests correctly.

##  Packages and data sets

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(error = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(DescTools)
library(dunn.test)
```

## Plan for today

- Tukey-Kramer tests with `TukeyHSD()`
- Bonferroni test for specific comparisons
- Dunnett's test with `DunnettTest()`
- Dunn's test with `dunn.test()`
- more fun with `ggplot2`
    + putting letters on plots with `geom_text()`
    + reordering plotted data with `factor()`

# But first... (a note about cheatsheets)


## Tukey-Kramer tests with `TukeyHSD()`

The Tukey-Kramer test runs all possible pairwise comparisons (all-ways comparisons) and produces a summary of which pairs are statistically different

```{r}
titanic <- read.csv("titanic.csv")
titanic.aov <- aov(fare ~ nationality, data = titanic)
anova(titanic.aov)
TukeyHSD(titanic.aov)
```

## Try on your own!

Try using `TukeyHSD()` on the air quality data set. Then compare the p-values you got from Tukey HSD to those you got from the Bonferroni test.

```{r}
air.aov <- aov(Ozone ~ factor(Month), data = airquality)
anova(air.aov)
```

## Tukey-Kramer tests with `TukeyHSD()`

```{r}
TukeyHSD(air.aov)
```



# Bonferroni Test

## Bonferroni test

- More powerful because it's on a subset of all possible comparisons.
- Uses pooled standard deviation from ANOVA, so better than doing individual t-tests.

$$
\alpha^* = \frac{\alpha}{n_{tests}}
$$
*OR*

$$
p^* = p \times n_{tests}
$$

## Implementing Bonferroni test in R

1. Do ANOVA, if significant...
2. Choose comparisons of interest
3. Use `pairwise.t.test()` with **no** p-value adjustment
4. Adjust p-values from comparisons of interest with `p.adjust()` (or using math) 

## Example: `airquality`

Does Ozone differ by Month?

```{r}
head(airquality)
```

## Step 1: ANOVA

(note: `Month` has to be manually converted to a factor)

```{r}
ozone.aov <- aov(Ozone ~ factor(Month), data = airquality)
anova(ozone.aov)
```

## Visualize Data

```{r echo=FALSE, warning=FALSE}
ggplot(airquality, aes(x = as.factor(Month), y = Ozone, fill = as.factor(Month))) +
  geom_boxplot() + labs(x = "Month", y = "Mean ozone (ppb)")
```

## Step 2: choose comparisons

Let's say you just want to know about how summer months compare to eachother

Comparisons of interest:

- June vs. July
- June vs. August
- July vs. August

## Step 3: `pairwise.t.test()`

```{r}
pairwise.t.test(airquality$Ozone, airquality$Month, p.adjust.method = "none")
```

## Step 4: Adjust p-values for comparisons of interest

- June vs. July: p = 0.01023
- June vs. August: p = 0.00831
- July vs. August: p = 0.91744

```{r}
ps <- c(0.01023, 0.00831, 0.91744)
p.adjust(ps, method = "bonf")
ps * length(ps) #does almost the same thing
```


## Dunnett's test with `DunnettTest()`

Dunnett's test runs pairwise comparisons of experimental groups against a control (many-to-one comparisons) and produces a summary of which pairs are statistically different

```{r}
library(DescTools)

DunnettTest(weight ~ feed, data = chickwts, control = "soybean")
```


## Dunnett's test with `DunnettTest()`

Try using `DunnettTest()` on the air quality data set!
```{r}
DunnettTest(Ozone ~ factor(Month), data = airquality, control = "9")
```


## Dunn's test with `dunn.test()`

Dunn's test is for non-parametric multiple comparisons (ANOVA vs. Kruskal-Wallis)

I remember last year there being some confusion about the p-value adjustment method.  I think the book/slides don't teach an optional adjustment method, but Dunn's method was equivalent to bonferroni I think. Maybe.

```{r}
library(dunn.test)
dunn.test(InsectSprays$count, InsectSprays$spray)
```

## Dunn's test with `dunn.test()`

```{r}
dunn.test(InsectSprays$count, InsectSprays$spray, method = "bonferroni")
```



# Plotting the results of multiple comparisons

## Step 1:

Stare at the test output, and decide what the letters are
```{r}
titanic.aov <- aov(fare ~ nationality, data = titanic)
anova(titanic.aov)
TukeyHSD(titanic.aov)
```

## Step 2:

Stare at the ggplot output, and decide where the letters should go
```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_boxplot()
```

## Step 2:

Stare at the ggplot output, and decide where the letters should go
```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_boxplot() + ylim(0,200)
```

## Step 3:

Make a dataframe with your label information

```{r}
label_table <- data.frame(x.valuez = c(1,2,3),
                          y.valuez = 175,
                          letterz = c("a","b","b"))
label_table
```

## Step 4:

Put it all together!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_boxplot() + ylim(0,200) + 
  geom_text(data = label_table, 
            aes(x = x.valuez,  y = y.valuez, label = letterz), 
            color = "red", size = 6)
```

## Step 5:

Have fun!!!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_jitter() +
  geom_text(data = label_table, 
            aes(x = x.valuez,  y = y.valuez, label = letterz), 
            color = "red", size = 6) 
```

## Step 5:

Have fun!!!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_jitter() +
  geom_text(data = label_table, 
            aes(x = x.valuez,  y = 400, label = letterz), 
            color = "red", size = 9) 
```

## Step 5:

Have fun!!!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_violin(fill = "dodgerblue") + ylim(0,200) +
  geom_text(data = label_table, 
            aes(x = x.valuez,  y = y.valuez, label = letterz), 
            color = "dodgerblue", size = 6) 
```

## Step 5:

Have fun!!!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_violin(fill = "dodgerblue") + ylim(0,200) +
  geom_text(data = label_table, 
            aes(x = x.valuez,  y = c(175, 163, 163), label = letterz), 
            color = "dodgerblue", size = 6) 
```

## Step 5:

Have fun!!!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_violin(fill = "dodgerblue") +
  geom_text(data = label_table, 
            aes(x = x.valuez - 0.1,  y = y.valuez, label = letterz), 
            color = "dodgerblue", size = 6) 
```

## Step 5:

Have fun!!!

```{r}
p <- ggplot(titanic, aes(x = nationality, y = fare))
p + geom_violin(fill = "dodgerblue") + ylim(0,200) +
  geom_text(data = label_table, 
            aes(x = x.valuez,  y = 20, label = letterz), 
            color = "white", size = 6) 
```