---
title: 'Multiple Comparisons: Packages and Functions'
author: "Eric R. Scott"
output:
  html_notebook: default
  pdf_document: default
---
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
```

# Multiple Comparisons

Built in to R:

- Tukey Tests (`TukeyHSD()`)
- Pairwise t-tests with a variety of p-value adjustments (e.g. Bonferoni) (`pairwise.t.test()`)
- Adjusting p-values for specific compairisons (`t.test()`, extract p-values, pass vector of p-values to `p.adjust()`)

Other multiple comparisons:

- Dunn's test---non-parametric mutliple comparisons following Kruskal Wallis (`dunn.test::dunn.test()`)
- Dunnet's test---compares mutiple groups to a control (many complicated options)

Options for Dunnet's test:

- `PMCMRplus::dunnettTest()`
- `multcomp::glht()` (https://stat.ethz.ch/pipermail/r-help/2015-June/429302.html)
- `DescTools::DunnettTest()`  <-- **This one is the best I think**

Intallation for all of these went smoothly for me.

# Tukey's HSD

We need a better example data set with fewer levels.  (like 3 or 4)

```{r}
chicks.aov <- aov(weight ~ feed, data = chickwts)
anova(chicks.aov)
TukeyHSD(chicks.aov)
```

# Bonferroni adjustments

## all pairwise tests

```{r}
pairwise.t.test(chickwts$weight, chickwts$feed, p.adj = "bonf")
```

## specific comparisons

The way I've done this is probably too complicated to be worth it.

```{r}
chickwts

# Spread data so you get each feed as a separate column.
chickwts.spread <- chickwts%>% 
  group_by(feed) %>% 
  mutate(ID = 1:n()) %>% # if you don't do this line, you get a "Duplicate identifiers" error
  spread(key = feed, value = weight) %>% 
  select(-ID) # removes the ID column that we used just to get it to spread correctly

comp1 <- t.test(chickwts.spread$horsebean, chickwts.spread$soybean)
#p = 0.0001559
comp2 <- t.test(chickwts.spread$horsebean, chickwts.spread$meatmeal)
#p=0.0001054

ps <- c(comp1$p.value, comp2$p.value)
my.comps <- data.frame(comparison = c("horsebean - soybean", "horsebean - meatmeal"),
                       p.vals = ps) %>% 
  mutate(p.adj = p.adjust(p.vals, "bonf"))
my.comps
```
you can also use p.adjust outside of mutate
```{r}
p.adjust(c(0.0001559, 0.0001054), "bonf")
```

# Dunn's Test

I remember last year there being some confusion about the p-value adjustment method.  I think the book/slides don't teach an optional adjustment method, but Dunn's method was equivalent to bonferroni I think. Maybe.

`altp = TRUE` is super important, I think.  From the help file:

"Reject Ho based on p ≤ α/2 (and in combination with p-value ordering for stepwise method options). If you prefer to work with p-values expressed as p-value = P(|Z| ≥ |z|) use the altp=TRUE option, and reject Ho based on p ≤ α (and in combination with p-value ordering for stepwise method options"

So by default, you look for values greater than alpha/2, but with `altp = TRUE` you look for p-values lower than alpha.

It's not clear to me if you need to adjust p-values or not.

```{r}
library(dunn.test)
dunn.test(InsectSprays$count, InsectSprays$spray, altp = TRUE)
# dunn.test(InsectSprays$count, InsectSprays$spray, altp = TRUE, method = "bonferroni")
```


# Dunnett's Test

Let's use chickwts and say soybean is the control.

## With `multcomp::glht()`

This one relies on factor levels to indicate which is the control.

```{r}
library(forcats) #for fool-proof manipulation of factor levels
data("chickwts")
```

1. Relevel so soybean is the control

```{r}
chickwts2 <- chickwts %>% 
  mutate(feed = fct_relevel(feed, "soybean")) #move soybean to the first level

levels(chickwts$feed)
levels(chickwts2$feed)
```

2. Run the `aov()` model

```{r}
chick.aov <- aov(weight ~ feed, data = chickwts2)
```

3. Do Dunnett's test with `glht()`

```{r}
library(multcomp)
thing <- glht(chick.aov, linfct = mcp(feed = "Dunnett"))
summary(thing)
```

**Pros**: nice output
**Cons**: Relies on factor levels to determine control. A lot of the code is "just because that's how you do it".  Not intuitive.  What is linfct? what is mcp?  Why does feed *equal* "Dunnett"?? Gotta run `summary()` to get p-values?


## With `DescTools::DunnettTest()`

```{r}
library(DescTools)
DunnettTest(weight ~ feed, data = chickwts, control = "soybean")
```

**Pros:** Looks just like `aov()` or `t.test()`.  Supply the control level directly.
**Cons:** Not as flexible as `glht()` (only does Dunnett's Test)


## With `PMCMRplus::dunnettTest()`

```{r}
library(PMCMRplus)
```

```{r}
dunnettTest(weight ~ feed, data = chickwts2) %>% summary()
```

**Pros:** Output is a data frame. Simpler than `glht`
**Cons:** Relies on factor levels for identifying control.


# Plotting Results of Multiple Comparisons

## Step 1:

Stare at the test output, decide what the letters are, make a vector.

```{r}
tooth.aov <- aov(len~factor(dose), data = ToothGrowth)
anova(tooth.aov)
TukeyHSD(tooth.aov)
```
Easy example!

a, b, c

```{r}
sig.letters <- c("a", "b", "c")
```

## Plotting Options

```{r message=FALSE, warning=FALSE}
library(ggplot2)
```

### Boxplot with letters

```{r}
p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len))
p +
  geom_boxplot() +
  stat_summary(geom = "text",
               label = sig.letters, #passed to geom_text()
               fun.y = max, #what function to use to summarize the data, ToothGrowth, to get y-values?
               position = position_nudge(y = 1.3)) #nudges the letters slightly above max(len)
```


### Jitter plot with means and SEM and letters nudged to the side?

```{r}
p + geom_jitter(width = 0.1) +
  stat_summary(geom = "point", fun.y = mean, color = "red", shape = "square", size = 3, position = position_nudge(x = 0.25)) +
  stat_summary(geom = "errorbar", fun.data = mean_se, color = "red", position = position_nudge(x = 0.25), width = 0.1) +
  stat_summary(geom = "text", label = sig.letters, fun.y = mean, position = position_nudge(x = 0.25, y = 2))
```


### Violin plots with letters??



