---
title: "Week 5 Live Coding Notebook"
author: "Eric R. Scott"
output: html_notebook
---

# Load necessary libraries
```{r}
library(dplyr)
library(ggplot2)
```


# Make a bar plot with error bars!

Let's use the chickwts dataset because it's familiar

```{r}
chickwts
```



## Make a summary table

mean, sd, SEM, 95%CI

```{r}
chick.summary <- chickwts %>% 
  group_by(feed) %>% 
  summarize(mean.wt = mean(weight),
            sd.wt = sd(weight),
            count = n(),                          # n() only works inside summarize() and never takes any arguments
            se.wt = sd(weight) / sqrt(count),     #I can use "count" because I created it in the previous line
            ci.wt = qt(0.025, count - 1, lower.tail = FALSE) * se.wt)
chick.summary
```

## Plot bars and error bars

```{r}
p <- 
  #set up the aesthetics for the plot using the summary data
  ggplot(chick.summary, aes(x = feed, y = mean.wt)) + 
  geom_col() +                                           
  geom_errorbar(aes(ymin = mean.wt - se.wt,
                    ymax = mean.wt + se.wt,
                    #`width` changes the width of the error bars
                    width = 0.25)) +                     
  labs(x = "Feed Type", y = "Chick Weight (g)")

p
```


# Add jittered points

Need to use *original* data for this

```{r}
#the `data` argument tells a geom_ to use a different dataset than the one in ggplot()
p + geom_jitter(data = chickwts,
                #you also have to re-define the aesthetics for the new data
                aes(x = feed, y = weight), alpha = 0.5, width = 0.25)
```



# Do the WHOLE THING with the original data

You can use `stat_summary()` to do it all with the original data.  This is a lot less coding, but it's less flexible and less intuitive.  Essentially, it does `summarize()` on your data and then passes it to a geom to plot it.  Specify how to summarize the data with `fun.y` or `fun.data` and specify the geom with `geom = `.

```{r}
ggplot(chickwts, aes(x = feed, y = weight)) +
  stat_summary(geom = "col", fun.y = mean) +
  stat_summary(geom = "errorbar", fun.data = mean_se, width = 0.25) +
  labs(x = "Feed Type", y = "Chick Weight (g)")
```


**Back to slides**

# Jitter plot with mean and error bars

```{r}
jp <- ggplot(chickwts, aes(x = feed, y = weight)) +
  geom_jitter(width = 0.2, alpha = 0.6) +       #alpha changes transparency
  geom_point(data = chick.summary, aes(x = feed, y = mean.wt),
             color = "blue",
             shape = 18,         #google "r plot shapes" to see all the possibiilties
             size = 3) +
  geom_errorbar(data = chick.summary,
                aes(x = feed,
                    ymin = mean.wt - ci.wt,
                    ymax = mean.wt + ci.wt),
                #without the next line, it tries to use the aesthetics from the original data
                inherit.aes = FALSE,  
                color = "blue", width = 0.25) +
  labs(x = "Feed Type", y = "Chick Weight (g)")
jp
```

# Oh, and if you want to angle the axis tick labels like I did in the slides...

```{r}
jp + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

