---
title: "Week 11: Correlation"
output: html_notebook
---

```{r packages and data}
library(ggplot2)
data(trees)
bees <- readRDS("BeeNectar.rds")
```
# What's a .rds file?
I did the same data wrangling steps we usually do with BeeData.csv--removing colony 22 and converting Colony to a factor--then saved it as a .rds file using saveRDS(). An .rds file can be used to save any R object and has the advantage that it contains information a .csv file can't.  For example, this file type remembers that `Colony` should be a factor, even though it contains numerical data.  Also, .rds files are typically smaller than .csv files. They can be useful if you have to do a lot of data wrangling---you can do the data wrangling in one R notebook, then save the result as a .rds and load it in a new notebook for analyses.  This just makes your analysis notebook easier to read.


# Plotting bivariate quantitative data
## Using `plot()` in base R
The `plot()` function can be used on a wide variety of objects and data types. Behind the scenes, it checks what type of data you have, and then applies the appropriate plotting method. Run `methods(plot)` if you want to see all the different types of methods it has and check their documentation if you're interested.  For today, when we run `plot()` it will be seeing that our data is continuous and bivariate and then running `plot.default()` to make a scatter plot.

```{r scatterplot}
#x,y interface
plot(trees$Girth, trees$Height)

#formula interface
plot(Height ~ Girth, data = trees)

#adding axis labels
plot(Height ~ Girth, data = trees, xlab = "Girth (in)", ylab = "Height (ft)")
```
For correlation, x and y don't matter, but for regression they do!
Checking assumption of bivariate normality should just be done visually (no formal tests for this).  Look for outliers, or non-linear patterns in the data.

## Using ggplot2
```{r ggplot scatterplot}
ggplot(trees, aes(x = Girth, y = Height)) +
  geom_point() +
  labs(x = "Girth (in)", y = "Height (ft)")
```

To deal with overplotting you could change the transparency of points with `alpha = ` (I also make the size bigger, and the background white in this example so the change in transparency is more obvious).


```{r adjust alpha}
ggplot(trees, aes(x = Girth, y = Height)) +
  geom_point(alpha = 0.4, size = 4) + #alpha sets transparency, size changes point size
  labs(x = "Girth (in)", y = "Height (ft)") +
  theme_bw() #just changes the appearance of the plot
```

# Pearson correlation
## Get a correlation coefficient
`cor()` only returns a correlation coefficient.  It's useful because it returns only a single thing compared to `cor.test()`.  This could come in handy if you wanted to calculate a bunch of correlation coefficients.  You can also use it on a data frame and get all pairwise correlations.
```{r}
cor(trees$Girth, trees$Height, method = "pearson")
cor(trees)
```

## Hypothesis testing
```{r}
cor.test(trees$Girth, trees$Height, method = "pearson")
```
The default method is pearson, but just for clarity, I've written it in the code.

What to report:

 - Test statistic r ("cor" in the R output)
 - N or df
 - p-value

# Spearman correlation
```{r}
cor(trees$Girth, trees$Height, method = "spearman")
cor.test(trees$Girth, trees$Height, method = "spearman")
```
What to report:

 - r_s ("rho" in the R output)
 - N or df
 - p-value

# Exercise
Test the hypothesis that ITS and nectar return are correlated using the bee data

## Plot to check assumptions
```{r}
bees
plot(bees$Nectar, bees$ITS)
```

Looks like a linear relationship.  Possibly some outliers with extreme values for nectar.  Data seems truncated at 0.  We could try a data transformation or go with a non-parametric method.  Square root transformed nectar returns looks better.

```{r}
plot(sqrt(bees$Nectar), bees$ITS, xlab = "sqrt(nectar return)", ylab = "ITS (mm)")
```

## Pearson product-moment correlation
Using sqrt transformed nectar data
```{r}
bee.cortest <- cor.test(sqrt(bees$Nectar), bees$ITS, method = "pearson")
bee.cortest
bee.cortest$estimate^2 #r-squared
```
Statistically significant, but pretty weak correlation


## Spearman rank correlation
Using non-parametric method on untransformed data is also a viable option

```{r}
cor.test(bees$Nectar, bees$ITS, method = "spearman")
```
Statistically significant relationship, but very weak correlation.

