---
title: "ABD Chapter 2 Plots re-done"
output:
  word_document: default
  html_notebook: default
---

```{r}
library(abd)
library(tidyverse)
```
```{r}
abdData()
#Locust data seems to be missing from abd package

locustData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f1_2locustSerotonin.csv"))
locustData
```

# Figure 2.1-2: Stripchart (aka Jitter Plot)
```{r}
ggplot(locustData, aes(x = treatmentTime, y = serotoninLevel)) +
  geom_jitter(width = 0.1)
```
Fancier version
```{r}
#find mean of each group
locustmeans <- locustData %>% 
  group_by(treatmentTime) %>% 
  summarise(mean = mean(serotoninLevel))

ggplot(locustData, aes(x = treatmentTime, y = serotoninLevel)) +
  geom_jitter(width = 0.1, color = "red") +
  labs(x = "Treatment time (hours)",
       y = "Seratonin (pmoles)") +
  geom_point(aes(x = treatmentTime, y = mean), data = locustmeans, size = 3, shape = 15) +
  theme_bw()
```

# Figure 2.1-3
```{r}
educationSpending <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f1_3EducationSpending.csv"))
educationSpending
```
```{r}
ggplot(educationSpending, aes(x = year, y = spendingPerStudent)) +
  geom_col()
```

fancier version:
```{r}
ggplot(educationSpending, aes(x = year, y = spendingPerStudent)) +
  geom_col(fill = "firebrick") +
  ylim(0, 8000) +
  labs(x = "Year", y = "Education spending ($ per student)") +
  theme_bw()
```

# Example 2.2A
```{r}
tigerData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e2aDeathsFromTigers.csv"))
tigerData
```
## Frequency table
```{r}
tigertable <- tigerData %>% 
  group_by(activity) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

tigertable 
```

## Barplot
using original data and geom_bar()
```{r}
ggplot(tigerData, aes(x = fct_infreq(activity))) +
  geom_bar()
```

Using summary table and geom_col
```{r}
ggplot(tigertable, aes(x = fct_reorder(activity, count, .desc = TRUE), y = count)) +
  geom_col()
```

Fancier version
```{r}
ggplot(tigertable, aes(x = fct_reorder(activity, count, .desc = TRUE), y = count)) +
  geom_col(fill = "firebrick") +
  labs(x = "Activity",
       y = "Frequency (number of people)") +
  theme_bw() +
  theme(axis.text.x.bottom = element_text(angle = 90))
```

# Example 2.2B
```{r}
birdAbundanceData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e2bDesertBirdAbundance.csv"))
birdAbundanceData
```

## Frequency table
I don't know how useful this really is, and it is not "tidy". Skip for now

## Histogram
```{r}
ggplot(birdAbundanceData, aes(x = abundance)) +
  geom_histogram()

#pick a better bin size or number of bins
ggplot(birdAbundanceData, aes(x = abundance)) +
  geom_histogram(binwidth = 50)
```

Fancier version:
```{r}
ggplot(birdAbundanceData, aes(x = abundance)) +
  geom_histogram(binwidth = 50, fill = "firebrick") +
  labs(x = "Abundance (No. Individuals)",
       y = "Frequency (No. Species)") +
  theme_bw()
```

# Figure 2.2-5.
```{r}
salmonSizeData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f2_5SalmonBodySize.csv"))
salmonSizeData
```

## Histograms with different numbers of bins
```{r}
p <- ggplot(salmonSizeData, aes(x = massKg))

p + geom_histogram()
p + geom_histogram(bins = 12)
p + geom_histogram(bins = 8)
```


## Bi-variate scatter plot
```{r}
ggplot(salmonSizeData, aes(x = lengthMm, y = massKg)) +
  geom_point(aes(color = as.factor(oceanAgeYears))) +
  scale_color_manual("Age", values = c("purple", "darkgreen")) +
  geom_smooth(method = "lm", se = FALSE, color = "black")

# colors()
```


# Example 2.3A.
```{r}
birdMalariaData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e3aBirdMalaria.csv"))
birdMalariaData
```

# Contingency table
These are inherently non-tidy.  I'm thinking we should wait to introduce until the very end when we tackle categorical data

# Grouped barplot
```{r}
ggplot(birdMalariaData, aes (x = treatment, fill = response)) +
  geom_bar()
```

## Mosaic Plot
```{r}
# #need to start with summary plot??
# birdMalariaData %>% 
#   group_by(treatment, response) %>% 
#     summarise(comb_cnt  = n()) %>%
#     mutate(height  = comb_cnt /sum(comb_cnt)) %>%
#     ungroup()......
# 
# tibble(treatment = c("Control", "Control", "Remove", "Remove"),
#        response = c("Yes", "No", "Yes", "No"),
#        height = c()).......
# 
# 
# birdtable$response
# ggplot(birdtable, aes(x = fct_inorder(response), y = freq, fill = treatment)) +
#    geom_bar(stat = "identity", position = "fill").......
```
ggmosaic appears to be broken.  Re-visit later
```{r}
library(ggmosaic)

#need to start with summary plot
birdtable <- birdMalariaData %>% 
  group_by(treatment, response) %>% 
  summarise(freq = n())

# ggplot(birdtable, aes(weight = freq, x = treatment, fill = response)) +
#   geom_mosaic()
```
# Example 2.3B.
```{r}
guppyFatherSonData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e3bGuppyFatherSonAttractiveness.csv"))
guppyFatherSonData
```

## Bi-variate scatter plot
```{r}
ggplot(guppyFatherSonData, aes(x = fatherOrnamentation, y = sonAttractiveness)) +
  geom_point()
```

fancier version:

```{r}
ggplot(guppyFatherSonData, aes(x = fatherOrnamentation, y = sonAttractiveness)) +
  geom_point(color = "firebrick", size = 3) +
  labs(x = "Father's ornamentation", y = "Son's attractiveness")
```


# Example 2.3C.
```{r}
#actualy exists in `abd` but not the same as what's on the web
# Hemoglobin
hemoglobinData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e3cHumanHemoglobinElevation.csv"))
hemoglobinData
```
## Summarise by population
```{r}
hemoglobinData %>%
  group_by(population) %>%
  summarise(count = n(),
            mean_hemoglobin = mean(hemoglobin))
```

## Stripchart/jitter plot
```{r}
ggplot(hemoglobinData, aes(x = population, y = hemoglobin)) +
  geom_jitter(width = 0.25)
```

Fancier version
```{r}
p <- ggplot(hemoglobinData, aes(x = population, y = hemoglobin)) +
    labs(x = "Male population",
       y = "Hemoglobin concentration (g/dL)")
#notice there is no geom_*.  We can add it to "p" later
  
p + geom_jitter(width = 0.25,
                alpha = 0.2,         #alpha adjusts transparency
                color = "firebrick",
                size = 3) +
  theme_bw()

```

## Boxplot
Using the same data
```{r}
p + geom_boxplot()
```

## Violin plot
Best of both worlds?

```{r}
p + geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), scale = "area", trim = FALSE)
```

## Faceted histograms
```{r}
ggplot(hemoglobinData, aes(x = hemoglobin)) +
  geom_histogram(bins = 20, fill = "firebrick") +
  facet_wrap("population", ncol = 1, scales = "free_y") +
  labs(x = "Hemoglobin concentration (g/dL)")
```

## ridge plot (density plot)
```{r}
library(ggridges)
ggplot(hemoglobinData, aes(x = hemoglobin, y = population, fill = population)) +
  geom_density_ridges()
```



# Example 2.4A.

```{r}
measlesData <- read_csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e4aMeaslesOutbreaks.csv"))
measlesData
```

## Line graph

```{r}
ggplot(measlesData, aes(x = yearByQuarter, y = confirmedCases)) +
  geom_line()
```

fancier version:
```{r}
ggplot(measlesData, aes(x = yearByQuarter, y = confirmedCases)) +
  geom_line() +
  geom_point(color = "blue") +
  labs(x = "Year", y = "Number of Cases", title = "Confirmed Measles Cases From 1995-2011") +
  theme_bw()
```

