---
title: "Sampling and Summarizing Data"
author: "Eric R. Scott, Avalon Owens"
output:
  ioslides_presentation:
  pdf_document: default
---

## Notes to self:
- students that decided to use R for repeated random samples in PS2 had a problem where the random samples were different when they knit to Word.  We could teach `set.seed()` or how to put R code inline.

## Plan for today
- Basic summary statistics 
- Reading in data with `read.csv()`
- Summarizing data with the `dplyr` package
- Sampling with `sample()` and `sample_n()`
- Making more cool plots with `ggplot2`

## Basic summary statistics (review)

```{r}
class_size <- c(10, 14, 37, 23, 42, 300, 55, 9, 278, 7)
mean(class_size)
round(mean(class_size), digits = 0)
```

## Basic summary statistics (new)

```{r}
median(class_size) #median
max(class_size) #maximum
min(class_size) #minimum
max(class_size) - min(class_size) #range
```

## Basic summary statistics (new)

```{r}
var(class_size) #variance
sd(class_size) #standard deviation
sd(class_size)^2 #also variance (I promise)
```

# But...

# What if you want to summarize a huge data frame into a nice table of results? 

# What if you want to calculate summary statistics for separate groups?
    
## Reading in data

- Export your Excel spreadsheet to .csv
- Put the .csv in same folder as your .Rmd
- Read in data with `read.csv("name.csv")`
- Assign this data frame to a variable
- Give it a short, descriptive name!

## Reading in data

```{r}
firefly <- read.csv("firefly_weight.csv")
head(firefly)
```

## Reading in data

```{r}
firefly <- na.omit(firefly)
head(firefly)
```

# Coding hack #1

## Named and unnamed arguments

- So far, we've been using named arguments for all functions:

```{r}
library(ggplot2)
firefly_plot <- ggplot(### <b>
data = firefly,
### </b>
                       aes(x = sex, y = weight))
```

- You don't have to name them (write `data = `) if you insert them into the function in the right order

```{r}
firefly_plot <- ggplot(firefly, aes(x = sex, y = weight))
```

- To figure out the right order, check the help file
    
## Summarizing data

- We're going to use the `dplyr` package to summarize data

`summarize(data, new_variable = function(variable))`

```{r message=FALSE, warning=FALSE}
library(dplyr)
summarize(firefly, mean_weight = mean(weight))
```

## Summarizing data

```{r}
summarize(firefly,
          mean = mean(weight),
          median = median(weight),
          range = max(weight) - min(weight),
          variance = var(weight),
          sd = sd(weight))
```

## Summarizing data by groups

- Sometimes we want summary statistics for groups of data

- Remember the faceted histograms we made last recitation?

- Use `group_by()` to add an *invisible* grouping variable

## Summarizing data by groups

```{r}
firefly_sex <- group_by(firefly, sex)

summarize(firefly_sex,
          mean = mean(weight),
          median = median(weight),
          range = max(weight) - min(weight),
          variance = var(weight),
          sd = sd(weight))
```

# Coding hack #2

## The "pipe" (`%>%`)

- Used in `dplyr` and many other Tidyverse packages
- macOS shortcut: command + shift + m
- Windows shortcut: control + shift + m
- Allows you to "chain" functions together

```{r}
firefly %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight))
```

## The "pipe" (`%>%`)

- Using the pipe makes your code more readable to humans!

```{r}
firefly %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight))
```

"Take the firefly dataset, group it by sex, then summarize it by taking the mean weight for each sex"

## The "pipe" (`%>%`)

- What it actually does is put whatever is on the left into the first argument of the function on the right

```{r eval = FALSE}
firefly %>% group_by(sex)
# is the same as
group_by(firefly, sex)
```

- Only works with functions that have `data` as first argument 
  + (This is all Tidyverse functions, but not all base R functions)

## Practice

- Make this code more readable using ` %>% `

```{r eval=FALSE}
iris_grouped <- group_by(iris, Species)

summarize(iris_grouped,
          mean = mean(Sepal.Length),
          median = median(Sepal.Length),
          range = max(Sepal.Length) - min(Sepal.Length),
          variance = var(Sepal.Length))
```

## Sampling

- Randomly sampling from a vector is quite simple

```{r}
class_size
sample(class_size, size = 3, replace = F)
sample(class_size, 3, F)
```

## Sampling

- Randomly sampling rows from a data frame is also easy!

```{r}
head(firefly)
sample_n(firefly, size = 3, replace = F)
sample_n(firefly, 3, F)
```

## Putting it all together

"Take the iris dataset, randomly sample 30 rows, group the data by species, then summarize the sample by taking the mean and standard deviation of petal length for each species... 

...and turn your random sample into a boxplot!"

## Putting it all together

```{r}
iris_sample <- iris %>% 
  sample_n(30, F)

iris_summary <- iris_sample %>% 
  group_by(Species) %>% 
  summarize(petal_length = mean(Petal.Length),
            sd_length = sd(Petal.Length))

iris_summary
```

## Putting it all together

```{r}
ggplot(iris_sample, aes(x = Species, y = Petal.Length)) + 
### <b>
  geom_boxplot() +
### </b>
  labs(x = "Iris species", y = "Petal length (cm)")
```

## Making more cool plots

```{r}
ggplot(iris_sample, aes(x = Species, y = Petal.Length)) + 
### <b>
  geom_violin() +
### </b>
  labs(x = "Iris species", y = "Petal length (cm)")
```

## Making more cool plots

```{r}
ggplot(iris_sample, aes(x = Species, y = Petal.Length)) + 
### <b>
  geom_point() +
### </b>
  labs(x = "Iris species", y = "Petal length (cm)")
```

## Making more cool plots

```{r}
ggplot(iris_sample, aes(x = Species, y = Petal.Length)) + 
### <b>
  geom_jitter(width = 0.25, height = 0) +
### </b>
  labs(x = "Iris species", y = "Petal length (cm)")
```

## Making more cool plots

```{r}
ggplot(iris_sample, aes(x = Sepal.Length, y = Petal.Length)) + 
### <b>
  geom_point() +
### </b>
  labs(x = "Sepal length (cm)", y = "Petal length (cm)")
```

# Coding hack #3

## Putting your name on your assignment

- What to do with the YAML header?